---
title: "2026 Summer Institute on<br>Longitudinal Data Analysis"
author: "Prepared by Kristine Marceau and Jennifer Hu"
date: "2026-07-06"
output:
  html_document:
    code-copy: true
    toc: yes
    css: purdue.css
    toc_float:
      collapsed: yes
      smooth_scroll: no
    toc_depth: 4
    includes:
      in_header: acumin-pro.html
      before_body: logo.html
  pdf_document:
    toc: yes
    toc_depth: 4
editor_options:
  markdown:
    wrap: 72
---


```{r setup-branding, include=FALSE}

brand_url <- "https://tdmize.github.io/data/Rmarkdown"
brand_files <- c("purdue.css", "acumin-pro.html", "logo.html")

for (f in brand_files) {
  try(
    download.file(
      paste(brand_url, f, sep = "/"),
      destfile = f,
      quiet = TRUE,
      mode = "wb"
    ),
    silent = TRUE
  )
}

```

# Overview

This file includes the data cleaning code for ABCD variables that can be
used as examples in the Methodology Center Summer Institute on
Longitudinal Data Analysis. For each section below, the code is
organized by topic/phenotype (level 1 header), a description of the
datasets and variables used according to ABCD documentation (level 2
header), the items and descriptions from each dataset are provided under
the dataset name (level 3 header). As applicable, the scored phenotypes
will be listed (level 4 headers). Finally, code for cleaning and
creating summary variables are also provided (level 2 headers).

**Please note that we did not correct any distributions/outliers - this
is up to each instructor.**

Data documentation can be found here:
<https://docs.abcdstudy.org/latest/documentation/curation/structure.html>

Load in packages and data (Hidden)

```{r, message=FALSE, warning=FALSE}

#clear working memory, if needed
rm(list = ls())

# Data manipulation
library(tidyverse)      # dplyr, ggplot2, readr, tidyr, tibble,
                        # purrr, stringr, forcats
# Reading ABCD data
library(arrow)          # read_parquet()

# Dates and times
library(lubridate)      # Date/time handling (visit dates)

# Descriptive statistics
library(psych)          # describe()
library(moments)        # skewness(), kurtosis()
                        # (used inside the plotting helper functions)

# Tables
library(kableExtra)     # kbl(), kable_minimal(),
                        # kable_styling(), collapse_rows(),
                        # scroll_box()

# Custom plotting functions
source("https://raw.githubusercontent.com/tdmize/Rfunctions/main/Viz_helper_functions.R")

```

# Codebook by Construct:

```{r construct-codebook, echo=FALSE, warning=FALSE, message=FALSE}

construct_codebook <- tribble(
  ~construct, ~indicator, ~description, ~waves_measured,

  "Identifiers", "<i>participant_id</i>", "Unique participant identifier.", "All waves (0-7)",
  "Metrics of Time", "<i>session_id</i>", "ABCD assessment session identifier.", "All waves (0-7)",
  "Metrics of Time", "<i>timepoint</i>", "Numeric timepoint extracted from session_id.", "All waves (0-7)",
  "Metrics of Time", "<i>ab_g_dyn__visit_dtt</i>", "Date and time at the start of the visit.", "All waves (0-7)",
  "Metrics of Time", "<i>ab_g_dyn__visit_age</i>", "Youth age at the start of the visit.", "All waves (0-7)",

  "Identifiers", "<i>ab_g_stc__design_id__fam</i>", "Family ID. Participants from the same family share a family ID.", "Baseline / time-invariant",
  "Identifiers", "<i>ab_g_stc__design_id__birth</i>", "Birth event ID. Participants born together share a birth event ID.", "Baseline / time-invariant",
  "Identifiers", "<i>ab_g_stc__design_famrel</i>", "Family relationship structure indicator.", "Baseline / time-invariant",
  "Identifiers", "<i>ab_g_dyn__design_site</i>", "Study Site. (1-21)", "All waves (0-7)",
   "Identifiers", "<i>zyg</i>", "Final zygosity/family relatedness variable. 1 = MZ twin; 2 = DZ twin; 3 = sibling.", "Baseline / time-invariant", 
  
  "Survey Weights", "<i>ab_g_stc__cohort_acsps</i>", "Imputed raked propensity weight", "Baseline",

  "Demographics and household", "<i>ab_g_stc__cohort_ethnrace__leg</i>", "Parent-reported child race/ethnicity. 1 = Hispanic; 2 = White; 3 = Black; 4 = Asian; 13 = Other.", "Baseline / time-invariant",
  "Demographics and household", "<i>ab_g_stc__cohort_race__nih</i>", "Child race based on NIH classification.", "Baseline / time-invariant",
  "Demographics and household", "<i>ab_g_stc__cohort_sex</i>", "Parent-reported sex assigned at birth. 1 = Male; 2 = Female.", "Baseline / time-invariant",
  "Demographics and household", "<i>ab_p_demo__relation_001</i>", "Relationship of the reporting caregiver to the participant.", "All waves (0-7)",
  "Demographics and household", "<i>ab_p_demo__marital__slf_001</i>", "Marital status of the reporting caregiver.", "All waves (0-7)",
  "Demographics and household", "<i>housepop</i>", "Number of people living at the youth's address.", "All waves (0-7)",
  "Demographics and household", "<i>multigen</i>", "Indicator for whether the youth lives in a multigenerational household.", "All waves (0-7)",
  "Demographics and household", "<i>ab_p_demo__child__time_001</i>", "Whether the child spends significant time in another household.", "All waves (0-7)",
  "Demographics and household", "<i>ab_p_demo__child__time_001__01</i>", "Amount of time per week the child spends in another household.", "All waves (0-7)",
  "Demographics and household", "<i>ab_p_demo__child__time_001__02</i>", "Best person to contact about the other household.", "All waves (0-7)",

  "Family environment and parenting", "<i>fc_y_fes__cohes_mean</i>", "Youth-reported Family Environment Scale cohesion mean score.", "Waves 4 and 7",
  "Family environment and parenting", "<i>fc_y_fes__confl_mean</i>", "Youth-reported Family Environment Scale conflict mean score.", "All waves (0-7)",
  "Family environment and parenting", "<i>ch_gen_know</i>", "Child-perceived parental monitoring/general knowledge, based on the youth-reported parental monitoring mean score.", "All waves (0-7)",
  "Family environment and parenting", "<i>par_su_rules</i>", "Parent-reported substance use rules composite. Higher scores indicate stricter and more strongly enforced rules regarding alcohol, tobacco, and marijuana.", "All waves (0-7)",

  "Peers", "<i>fc_y_rpi_mean</i>", "Youth-reported Resistance to Peer Influence mean score. Higher scores indicate greater resistance to peer influence.", "Waves 4-6",
  "Peers", "<i>positive_peer_mean</i>", "Mean of positive peer characteristic items, including friends being athletes, attending religious/spiritual services, and being excellent students.", "Waves 2-7",
  "Peers", "<i>negative_peer_mean</i>", "Mean of negative peer behavior items, including friends skipping school, being suspended, and shoplifting.", "Waves 2-7",

  "Mental health", "<i>mh_p_cbcl__synd__ext_sum</i>", "Parent-reported CBCL externalizing syndrome scale sum.", "All waves (0-7)",
  "Mental health", "<i>mh_p_cbcl__synd__int_sum</i>", "Parent-reported CBCL internalizing syndrome scale sum.", "All waves (0-7)",
  "Mental health", "<i>mh_p_cbcl__synd__attn_sum</i>", "Parent-reported CBCL attention problems syndrome scale sum.", "All waves (0-7)",
  "Mental health", "<i>mh_y_bpm__ext_sum</i>", "Youth-reported Brief Problem Monitor externalizing sum score.", "Waves 1-6",
  "Mental health", "<i>mh_y_bpm__int_sum</i>", "Youth-reported Brief Problem Monitor internalizing sum score.", "Waves 1-6",
  "Mental health", "<i>mh_y_pai_sum</i>", "Youth-reported NIH Toolbox Positive Affect Items sum score.", "Waves 1, 3, and 5",

  "Substance use risk", "<i>su_y_sui__hrd__alc_001</i>", "Heard of alcohol, such as beer, wine, or liquor. 0 = No; 1 = Yes.", "Full data at wave 0; limited data at waves 1-2",
  "Substance use risk", "<i>su_y_sui__hrd__nic_001</i>", "Heard of tobacco products. 0 = No; 1 = Yes.", "Full data at wave 0; limited data at waves 1-2",
  "Substance use risk", "<i>su_y_sui__hrd__mj_001</i>", "Heard of marijuana, weed, pot, blunts, dabs, marijuana drinks, or food with marijuana. 0 = No; 1 = Yes.", "Full data at wave 0; limited data at waves 1-2",
  "Substance use risk", "<i>su_y_itu__alc_001</i>-<i>su_y_itu__alc_003</i>", "Alcohol curiosity, intention to try soon, and willingness to try if offered by a best friend.", "All waves (0-7)",
  "Substance use risk", "<i>su_y_itu__nic_001</i>-<i>su_y_itu__nic_003</i>", "Tobacco curiosity, intention to try soon, and willingness to try if offered by a best friend.", "All waves (0-7)",
  "Substance use risk", "<i>su_y_itu__mj_001</i>-<i>su_y_itu__mj_003</i>", "Marijuana curiosity, intention to try soon, and willingness to try if offered by a best friend.", "All waves (0-7)",

  "Substance use", "<i>AUC</i>", "Alcohol uptake continuum. Higher scores indicate greater alcohol use involvement.", "All waves (0-7)",
  "Substance use", "<i>AU_init</i>", "Alcohol use initiation indicator. 0 = no initiation; 1 = alcohol uptake score greater than or equal to 1.", "All waves (0-7)",
  "Substance use", "<i>TUC</i>", "Tobacco uptake continuum. Higher scores indicate greater tobacco/nicotine use involvement.", "All waves (0-7)",
  "Substance use", "<i>TU_init</i>", "Tobacco use initiation indicator. 0 = no initiation; 1 = tobacco uptake score greater than or equal to 1.", "All waves (0-7)",
  "Substance use", "<i>MUC</i>", "Marijuana uptake continuum. Higher scores indicate greater marijuana use involvement.", "All waves (0-7)",
  "Substance use", "<i>MU_init</i>", "Marijuana use initiation indicator. 0 = no initiation; 1 = marijuana uptake score greater than or equal to 1.", "All waves (0-7)",
  "Substance use", "<i>SUC</i>", "Overall substance use continuum/involvement score, calculated from alcohol, tobacco, and marijuana uptake continua.", "All waves (0-7)",
  "Substance use", "<i>SU_init</i>", "Any substance use initiation indicator. 0 = no substance use initiation; 1 = initiation of alcohol, tobacco, or marijuana use.", "All waves (0-7)",

  "Substance use", "<i>fulldrink</i>", "Combined baseline/longitudinal full drink indicator from su_y_sui__use__alc_001 and su_y_sui__use__alc_001__l.", "All waves (0-7)",
  "Substance use", "<i>sipalc</i>", "Combined baseline/longitudinal sip of alcohol indicator from su_y_lowuse__isip_002 and su_y_lowuse__isip_002__l.", "All waves (0-7)",
  "Substance use", "<i>puffnic</i>", "Combined baseline/longitudinal puff nicotine indicator from su_y_lowuse__nic_001 and su_y_lowuse__nic_001__l.", "All waves (0-7)",
  "Substance use", "<i>fullnic</i>", "Combined baseline/longitudinal full cigarette use indicator from su_y_sui__use__nic__cig_001 and su_y_sui__use__nic__cig_001__l.", "All waves (0-7)",
  "Substance use", "<i>puffmj</i>", "Combined baseline/longitudinal marijuana puff/use indicator from su_y_sui__use__mj__puff_001 and su_y_sui__use__mj__puff_001__l.", "All waves (0-7)",
  "Substance use", "<i>fullmj</i>", "Combined baseline/longitudinal smoked marijuana indicator from su_y_sui__use__mj__smoke_001 and su_y_sui__use__mj__smoke_001__l.", "All waves (0-7)",

  "Substance use", "<i>su_y_tlfb__alc__binge_ud</i>; <i>su_y_tlfb__alc_ud</i>; <i>su_y_tlfb__alc_mean</i>", "Alcohol binge episode use days, alcohol use days, and average dose.", "All waves (0-7)",
  "Substance use", "<i>su_y_tlfb__nic__cig_mean</i>; <i>su_y_tlfb__nic__cig_ud</i>", "Nicotine cigarette average dose and number of use days.", "All waves (0-7)",
  "Substance use", "<i>su_y_tlfb__mj__smoke_mean</i>; <i>su_y_tlfb__mj__smoke_ud</i>", "Average smoked marijuana on use days and number of smoked marijuana use days.", "All waves (0-7)",
  "Substance use", "<i>su_y_alcsre__6mo_count</i>; <i>su_y_alcsre__hvy_count</i>", "Subjective experiences from drinking alcohol in the past 6 months and during heaviest drinking period.", "All waves (0-7)",
  "Substance use", "<i>su_y_nicsre__cig_sum</i>", "Total intensity of positive and negative effects of first cigarette use.", "All waves (0-7)",
  "Substance use", "<i>su_y_mjsre_sum</i>", "Total number and severity of subjective responses for first marijuana use episode.", "All waves (0-7)",
  
    # Neighborhood context
  "Neighborhood context", "<i>le_l_adi__addr1__income_med</i>", "Median family income from the Area Deprivation Index, a neighborhood-level composite measure of socioeconomic disadvantage. Higher values indicate greater neighborhood deprivation.", "Baseline / time-invariant",
  "Neighborhood context", "<i>le_l_adi__addr1__mortgage_med</i>", "Median monthly mortgage from the Area Deprivation Index, a neighborhood-level composite measure of socioeconomic disadvantage. Higher values indicate greater neighborhood deprivation.", "Baseline / time-invariant",
  "Neighborhood context", "<i>le_l_svi__addr1__17younger_prcnt</i>", "Percent population estimate of persons 17 and younger from the Social Vulnerability Index, a neighborhood-level composite measure of community vulnerability.", "Baseline / time-invariant",
  "Neighborhood context", "<i>le_l_densalc__addr1__alc_density__2016</i>", "Estimate of alcohol outlet density based on Census 2016 ZIP Code Business Patterns and NAICS establishments where alcohol could be purchased and/or consumed.", "Baseline / time-invariant"
)

construct_codebook %>%
  arrange(
    factor(
      construct,
      levels = c(
        "Identifiers",
        "Metrics of Time",
        "Demographics and household",
        "Family environment and parenting",
        "Peers",
        "Mental health",
        "Substance use risk",
        "Substance use"
      )
    ),
    indicator
  ) %>%
  kable(
    format = "html",
    col.names = c("Construct", "Variable Name", "Variable Description", "Waves Measured"),
    escape = FALSE
  ) %>%
  kable_styling(
    bootstrap_options = c("hover", "condensed"),
    full_width = TRUE
  ) %>%
  collapse_rows(
    columns = 1,
    valign = "top"
  )
```

## Note on Session_id

In ABCD 7.0, session_id includes timepoints with S (screener), A (annual
assessment), and M (midpoint assessment). We maintain only sessions from
the annual assessment (i.e., ses-#A).

## Note on Viz Helper Functions

Available visualization helper functions:

**plot_nominal()**

Creates a bar chart for a categorical (nominal) variable and displays
counts and percentages for each category. Categories are ordered by
frequency and plotted using the project color palette.

*Key arguments:* data = data frame var = categorical variable to plot
title = optional plot title xlab = optional x-axis label ylab = optional
y-axis label (default = "Count") show_percent = whether to display
percentages above bars (default = TRUE)

**plot_distribution()**

Creates a histogram for a single numeric variable and displays N, mean,
SD, skewness, and kurtosis, along with reference lines at the mean, ±1
SD, and ±3 SD.

*Key arguments:* data = data frame var = numeric variable to plot title
= optional plot title xlab = optional x-axis label annotate_x = optional
x-position for summary statistics

**plot_distribution_by_time()**

Creates faceted histograms by timepoint (or another grouping variable),
with group-specific summary statistics and reference lines calculated
separately within each facet.

*Key arguments:* data = data frame var = numeric variable to plot
time_var = grouping/facet variable title = optional plot title xlab =
optional x-axis label annotate_x = optional x-position for summary
statistics

**plot_longitudinal_subset()**

Creates a longitudinal "spaghetti plot" for a random subset of
participants, families, or other repeated-measures units, with an
overall LOESS trend line.

*Key arguments:* data = long-format data frame var = outcome variable
time_var = time variable id_var = participant/family ID sample_n =
number of IDs to sample (default = 100) seed = random seed for
reproducibility title, xlab, ylab = optional labels

**plot_cormat_heatmap()**

Creates an upper-triangle Pearson correlation heatmap for a set of
numeric variables, displaying the correlation coefficient in each cell.

*Key arguments:* wide_subset = data frame containing variables of
interest

Load in datasets

```{r include=FALSE}

#Load in datasets

## Filepaths
## For Kristine
filepath <- "C:/Users/kmarcea/Box/[L2 Sensitive Ext] Marceau BDL/"
## For Jennifer
#filepath <- "C:/Users/admin/Box/"

# Demographics, identifiers
demo <- read_parquet(file=paste0(filepath,"ABCD/Data/abcd-data-release-7.0/rawdata/phenotype/ab_p_demo.parquet"), as_data_frame = TRUE)
stc <- read_parquet(file=paste0(filepath,"ABCD/Data/abcd-data-release-7.0/rawdata/phenotype/ab_g_stc.parquet"), as_data_frame = TRUE)
dyn <- read_parquet(file=paste0(filepath,"ABCD/Data/abcd-data-release-7.0/rawdata/phenotype/ab_g_dyn.parquet"), as_data_frame = TRUE)
zygrat <- read_parquet(file=paste0(filepath,"ABCD/Data/abcd-data-release-7.0/rawdata/phenotype/gn_e_zygrat.parquet"), as_data_frame = TRUE)
genrel <- read_parquet(file=paste0(filepath,"ABCD/Data/abcd-data-release-7.0/rawdata/phenotype/gn_y_genrel.parquet"), as_data_frame = TRUE)

# Emotional/Behavioral Risk for SU
bpm <- read_parquet(file=paste0(filepath,"ABCD/Data/abcd-data-release-7.0/rawdata/phenotype/mh_y_bpm.parquet"), as_data_frame = TRUE)
cbcl <- read_parquet(file=paste0(filepath,"ABCD/Data/abcd-data-release-7.0/rawdata/phenotype/mh_p_cbcl.parquet"), as_data_frame = TRUE)

# Positive Affects
pai <- read_parquet(file=paste0(filepath,"ABCD/Data/abcd-data-release-7.0/rawdata/phenotype/mh_y_pai.parquet"), as_data_frame = TRUE)

# Substance Use
sui <- read_parquet(file=paste0(filepath,"ABCD/Data/abcd-data-release-7.0/rawdata/phenotype/su_y_sui.parquet"), as_data_frame = TRUE)
lowuse <- read_parquet(file=paste0(filepath,"ABCD/Data/abcd-data-release-7.0/rawdata/phenotype/su_y_lowuse.parquet"), as_data_frame = TRUE)
itu <- read_parquet(file=paste0(filepath,"ABCD/Data/abcd-data-release-7.0/rawdata/phenotype/su_y_itu.parquet"), as_data_frame = TRUE)
tlfb <- read_parquet(file=paste0(filepath,"ABCD/Data/abcd-data-release-7.0/rawdata/phenotype/su_y_tlfb.parquet"), as_data_frame = TRUE)
alcsre <- read_parquet(file=paste0(filepath,"ABCD/Data/abcd-data-release-7.0/rawdata/phenotype/su_y_alcsre.parquet"), as_data_frame = TRUE)
nicsre <- read_parquet(file=paste0(filepath,"ABCD/Data/abcd-data-release-7.0/rawdata/phenotype/su_y_nicsre.parquet"), as_data_frame = TRUE)
mjsre <- read_parquet(file=paste0(filepath,"ABCD/Data/abcd-data-release-7.0/rawdata/phenotype/su_y_mjsre.parquet"), as_data_frame = TRUE)

# Env/Family Vars
pm <- read_parquet(file=paste0(filepath,"ABCD/Data/abcd-data-release-7.0/rawdata/phenotype/fc_y_pm.parquet"), as_data_frame = TRUE)
surule <- read_parquet(file=paste0(filepath,"ABCD/Data/abcd-data-release-7.0/rawdata/phenotype/su_p_rule.parquet"), as_data_frame = TRUE)

fes <- read_parquet(file=paste0(filepath,"ABCD/Data/abcd-data-release-7.0/rawdata/phenotype/fc_y_fes.parquet"), as_data_frame = TRUE)

pbp <- read_parquet(file=paste0(filepath,"ABCD/Data/abcd-data-release-7.0/rawdata/phenotype/fc_y_pbp.parquet"), as_data_frame = TRUE)

rpi <- read_parquet(file=paste0(filepath,"ABCD/Data/abcd-data-release-7.0/rawdata/phenotype/fc_y_rpi.parquet"), as_data_frame = TRUE)

# Linked external data
adi <- read_parquet(file=paste0(filepath,"ABCD/Data/abcd-data-release-7.0/rawdata/phenotype/le_l_adi.parquet"), as_data_frame = TRUE)
svi <- read_parquet(file=paste0(filepath,"ABCD/Data/abcd-data-release-7.0/rawdata/phenotype/le_l_svi.parquet"), as_data_frame = TRUE)
densalc <- read_parquet(file = paste0(filepath, "ABCD/Data/abcd-data-release-7.0/rawdata/phenotype/le_l_densalc.parquet"),as_data_frame = TRUE)



```

# Demographic Variables

## Dataset(s) & Variables Used Note:

This section creates demographic, family structure, and visit-level
variables from ABCD parent-report and administrative datasets. These
variables include caregiver relationship to the child, caregiver marital
status, household size and composition, the child’s time in another
household, child race/ethnicity and sex assigned at birth, family and
birth-event identifiers, family relationship structure, youth age at
visit, and visit date/time.

**ab_p_demo**

\*ab_p_demo\_\_relation_001\* (formerly demo_prim & demo_prim_l):
Relationship to the participant: You are the... 1 = Child's biological
mother, 2 = Child's biological father, 3 = adoptive parent, 4 = Child's
custodial parent, 5 = other

\*ab_p_demo\_\_marital\_\_slf_001\* (formerly demo_prnt_marital_v2 &
demo_prnt_marital_v2_l) 1 = Married, 2 = Widowed, 3 = Divorced, 4 =
Separated, 5 = Never married, 6 = Living with partner, 777 = decline to
answer

\*ab_p_demo\_\_roster_001\* (formerly demo_roster_v2 &
demo_roster_v2_l): How many people are living at your address?

\*ab_p_demo\_\_roster_002 through ab_p_demo\_\_roster_015\* (formerly
fam_roster_2c_v2 through fam_roster_15c_v2 & fam_roster_2c_v2_l through
fam_roster_15c_v2_l): Person #2 through Person #15 who lives at your
address, relationship to you. 1 = Spouse; 2 = child-in-law; 3 =
Biological child; 4 = Other relative; 5 = Adopted child; 6 = Roomer or
boarder; 7 = Stepchild; 8 = Housemate or roommate; 9 = sibling; 10 =
Unmarried partner; 11 = parent; 12 = Foster child; 13 = Grandchild; 14 =
Other nonrelative; 15 = Parent-in-law

\*ab_p_demo\_\_child\_\_time_001\* (formerly demo_child_time_v2 &
demo_child_time_v2_l): Is there another household in which the child
spends a significant amount of time? 1 = Yes; 0 = No; 777 = Declined to
answer

\*ab_p_demo\_\_child\_\_time_001\_\_01\* (formerly demo_child_time2_v2 &
demo_child_time2_v2_l): About how much time per week does the child
spend at this other household?

\*ab_p_demo\_\_child\_\_time_001\_\_02\* (formerly demo_child_time3_v2 &
demo_child_time3_v2_l): Who is the best person to contact about that
household? 1 = other biological parent; 2 = step mother; 3 = step
father; 4 = grandmother; 5 = grandfather; 6 = aunt; 7 = uncle; 8 = other

**ab_g_stc**

\*ab_g_stc\_\_cohort_acsps\* Cohort description: Imputed raked propensity weight - The raked propensity weight merges the ACS and ABCD baseline data (with missing data imputed), estimates the propensity model, computes and scales/trims the propensity weights and finally rakes the scaled weights to final ACS control totals by age, sex and race/ethnicity [Based on baseline demographics data]

\*ab_g_stc\_\_cohort_ethnrace\_\_leg\* (formerly race_ethnicity in
abcd_p_demo): Parent reported race for child. 1 = Hispanic; 2 = White; 3
= Black; 4 = Asian; 13 = Other

\*ab_g_stc\_\_cohort_race\_\_nih\*: Race (NIH classification reporting 7
levels) [Based on baseline response; missingness filled in from
longitudinal responses]

\*ab_g_stc\_\_cohort_sex\* (formerly demo_sex_v2 in abcd_p_demo): Parent
reported sex assigned at birth. 1=Male, 2=female

\*ab_g_stc\_\_design_id\_\_fam\* (formerly rel_family_id):
Design/nesting: Family ID - Participants belonging to the same family
share a family ID (derived based on caregiver information)

\*ab_g_stc\_\_design_id\_\_birth\* (formerly rel_birth_id):
Design/nesting: Birth event ID - Participants from the same family who
were born together share a birth event ID (derived based on unique
birthdates)

\*ab_g_stc\_\_design_famrel\* (formerly rel_relationship):
Design/nesting: Relationship of the participant in their family

**ab_g_dyn**

\*ab_g_dyn\_\_visit_age\* (formerly demo_brthdat_v2 &
demo_brthdat_v2_l): Youth's age at the start of the event

\*ab_g_dyn\_\_visit_dtt\* (formerly interview_date): Visit information:
Date and time at the start of the visit

\*ab_g_dyn__design_site*: Study Site

## Data Cleaning & Variable Creation

```{r demo-vars, message=FALSE}


#Subset demo dataset to variables of interest
s.demo <- c("participant_id","session_id", "ab_p_demo__relation_001","ab_p_demo__marital__slf_001", "ab_p_demo__roster_001", "ab_p_demo__roster_002", "ab_p_demo__roster_003", "ab_p_demo__roster_004", "ab_p_demo__roster_005", "ab_p_demo__roster_006", "ab_p_demo__roster_007", "ab_p_demo__roster_008", "ab_p_demo__roster_009", "ab_p_demo__roster_010", "ab_p_demo__roster_011", "ab_p_demo__roster_012", "ab_p_demo__roster_013", "ab_p_demo__roster_014", "ab_p_demo__roster_015", "ab_p_demo__child__time_001", "ab_p_demo__child__time_001__01", "ab_p_demo__child__time_001__02")
demo_subset <- demo[s.demo]

#Subset dyn dataset to variables of interest
s.dyn <- c("participant_id", "session_id", "ab_g_dyn__visit_age", "ab_g_dyn__visit_dtt","ab_g_dyn__design_site")
dyn_subset <- dyn[s.dyn]

#Subset stc dataset to variables of interest
s.stc <- c("participant_id", "ab_g_stc__cohort_ethnrace__leg", "ab_g_stc__cohort_race__nih","ab_g_stc__cohort_acsps", "ab_g_stc__cohort_sex", "ab_g_stc__design_id__fam", "ab_g_stc__design_id__birth", "ab_g_stc__design_famrel")
stc_subset <- stc[s.stc]


#merge select var datasets for demographics of demo_subset, stc_subset, & dyn_subset
demo_datavars <- list(demo_subset, dyn_subset)
demo_merged <- Reduce(function(x, y) merge(x, y, by = c("participant_id","session_id"), all = TRUE), demo_datavars)

#As stc doesn't have a session_id var, merge only by participant_id
#merge dataframes
demo_m <- merge(demo_merged, stc_subset, by=c("participant_id"), all = TRUE)

#create multigenerational home variable If parent (11) or grandchild (13) or parent-in-law (15) selected, then 1 (i.e., parent/guardian and child live with ), all else 0. 
demo_m <- demo_m %>% 
  mutate(multigen = case_when(
    (ab_p_demo__roster_002 == 11 | ab_p_demo__roster_002 == 13 | ab_p_demo__roster_002 == 15) ~ 1, 
    (ab_p_demo__roster_003 == 11 | ab_p_demo__roster_003 == 13 | ab_p_demo__roster_003 == 15) ~ 1,
    (ab_p_demo__roster_004 == 11 | ab_p_demo__roster_004 == 13 | ab_p_demo__roster_004 == 15) ~ 1,
    (ab_p_demo__roster_005 == 11 | ab_p_demo__roster_005 == 13 | ab_p_demo__roster_005 == 15) ~ 1,
    (ab_p_demo__roster_006 == 11 | ab_p_demo__roster_006 == 13 | ab_p_demo__roster_006 == 15) ~ 1,
    (ab_p_demo__roster_007 == 11 | ab_p_demo__roster_007 == 13 | ab_p_demo__roster_007 == 15) ~ 1,
    (ab_p_demo__roster_008 == 11 | ab_p_demo__roster_008 == 13 | ab_p_demo__roster_008 == 15) ~ 1,
    (ab_p_demo__roster_009 == 11 | ab_p_demo__roster_009 == 13 | ab_p_demo__roster_009 == 15) ~ 1,
    (ab_p_demo__roster_010 == 11 | ab_p_demo__roster_010 == 13 | ab_p_demo__roster_010 == 15) ~ 1,
    (ab_p_demo__roster_011 == 11 | ab_p_demo__roster_011 == 13 | ab_p_demo__roster_011 == 15) ~ 1,
    (ab_p_demo__roster_012 == 11 | ab_p_demo__roster_012 == 13 | ab_p_demo__roster_012 == 15) ~ 1,
    (ab_p_demo__roster_013 == 11 | ab_p_demo__roster_013 == 13 | ab_p_demo__roster_013 == 15) ~ 1,
    (ab_p_demo__roster_014 == 11 | ab_p_demo__roster_014 == 13 | ab_p_demo__roster_014 == 15) ~ 1,
    (ab_p_demo__roster_015 == 11 | ab_p_demo__roster_015 == 13 | ab_p_demo__roster_015 == 15) ~ 1,
    TRUE ~ 0))

demo_m <- demo_m %>% 
  mutate(housepop = ab_p_demo__roster_001)

#Drop roster variables that are not longer needed
s.demor <- c("participant_id", "session_id", "ab_g_dyn__visit_dtt", "ab_g_dyn__visit_age", "ab_p_demo__relation_001", "ab_p_demo__marital__slf_001", "multigen", "ab_p_demo__child__time_001", "ab_p_demo__child__time_001__01", "ab_p_demo__child__time_001__02", "ab_g_stc__cohort_ethnrace__leg", "ab_g_stc__cohort_race__nih", "ab_g_stc__cohort_sex", "ab_g_stc__design_id__fam", "ab_g_stc__design_id__birth", "ab_g_stc__design_famrel","housepop","ab_g_dyn__design_site")
demo_mf <- demo_m[s.demor]

```

## Descriptive Statistics

```{r, echo=FALSE, warning=FALSE, message=FALSE}

#Make a wide dataset
demowide = demo_mf %>%
  pivot_wider(
    id_cols = "participant_id", 
    names_from = "session_id", 
    values_from = c("ab_g_dyn__visit_age", "ab_p_demo__relation_001", "ab_p_demo__marital__slf_001", "multigen","ab_g_stc__cohort_ethnrace__leg",  "ab_g_stc__cohort_sex","ab_g_dyn__design_site"))

kbl(psych::describe(demowide), caption = "Demographic Summary Statistics") %>% 
  kable_minimal(c("striped", "hover","condensed","responsive")) %>%
  scroll_box(width = "100%")
```

## Viz: Frequency of Participants Sex At Each Timepoint

```{r, echo=FALSE, warning=FALSE, message=FALSE}

subset_demo.sex<-subset(demo_mf, !is.na(ab_g_stc__cohort_sex))

#demo_sex_v2: Parent reported sex assigned at birth. 1=Male, 2=female, 3=intersex-male, 4=intersex-female

ggplot(subset_demo.sex, aes(x = session_id, fill = factor(ab_g_stc__cohort_sex))) + 
  geom_bar(color = "black", position = "dodge") +
  scale_fill_manual(values = c("black","#CFB991", "#DAAA00", "#EBD99F"), labels = c("Male", "Female", "Intersex Male", "Intersex Female")) +
  labs(x = "Timepoint", y = "Frequency",
       fill = "Groups")+
  theme_minimal()

plot_nominal(
  data = subset_demo.sex,
  var = ab_g_stc__cohort_sex,
  title = "Sex Distributions Across All Waves",
  xlab = "Sex"
)

```

## Viz: Frequency of Participants Within Race-Ethnic Groups At Each Timepoint

```{r, echo=FALSE, warning=FALSE, message=FALSE}
subset_demo.race<-subset(demo_mf, !is.na(ab_g_stc__cohort_ethnrace__leg))

#race_ethnicity: 1 = Hispanic; 2 = White; 3 = Black; 4 = Asian; 13 = Other

subset_demo.race <- subset_demo.race %>%
  mutate(
    race_ethn_label = factor(
      ab_g_stc__cohort_ethnrace__leg,
      levels = c(1, 2, 3, 4, 13),
      labels = c("Hispanic", "White", "Black", "Asian", "Other")
    )
  )

ggplot(subset_demo.race, aes(x = session_id, fill = race_ethn_label)) + 
  geom_bar(color = "black", position = "dodge") +
  scale_fill_manual(
    values = c("Hispanic" = "black","White" = "#8E6F3E","Black" = "#DAAA00","Asian" = "#DDB945", "Other" = "#EBD99F")) +labs(x = "Timepoint",y = "Frequency",fill = "Groups") +
  theme_minimal()

plot_nominal(
  data = subset_demo.race,
  var = race_ethn_label,
  title = "Race/Ethnicity Distributions Across All Waves",
  xlab = "Race/Ethnicity"
)


```

## Viz: Frequency of Reporting Parent's Marital Status At Each Timepoint

```{r, echo=FALSE, warning=FALSE, message=FALSE}
subset_demo.marital<-subset(demo_mf, !is.na(ab_p_demo__marital__slf_001))
subset_demo.marital <- subset(subset_demo.marital, ab_p_demo__marital__slf_001 != 777)

subset_demo.marital$marital_label <- factor(
  subset_demo.marital$ab_p_demo__marital__slf_001,
  levels = c(1, 2, 3, 4, 5, 6),
  labels = c("Married", "Widowed", "Divorced", "Separated", 
             "Never married", "Living with partner")
)

#marital: marital status of the reporting parent. 1 = Married; 2 = Widowed ; 3 = Divorced; 4 = Separated; 5 = Never married; 6 = Living with partner; 777 = Decline to answer

ggplot(subset_demo.marital, aes(x = session_id, fill = factor(ab_p_demo__marital__slf_001))) + 
  geom_bar(color = "black", position = "dodge") +
  scale_fill_manual(values = c("black","#8E6F3E", "#DAAA00", "#DDB945", "#CFB991", "#EBD99F"), labels = c("Married", "Widowed", "Divorced", "Separated", "Never married", "Living with partner")) +
  labs(x = "Timepoint", y = "Frequency",
       fill = "Groups")+
  theme_minimal()


plot_nominal(
  data = subset_demo.marital,
  var = marital_label,
  title = "Caregiver Marital Status Distributions Across All Waves",
  xlab = "Marital Status"
)
```

## Viz: Frequency of Participants Living Within a Multigenerational Home

```{r, echo=FALSE, warning=FALSE, message=FALSE}
subset_demo.multigen<-subset(demo_mf, !is.na(multigen))

#multigen: Created variable that denotes child lives in a multigenerational home (1) or not (0).


subset_demo.multigen$multigen_label <- factor(
  subset_demo.multigen$multigen,
  levels = c(0, 1),
  labels = c(
    "Not Living in a Multigenerational Home",
    "Living in a Multigenerational Home"
  )
)

ggplot(subset_demo.multigen, aes(x = session_id, fill = multigen_label)) + 
  geom_bar(color = "black", position = "dodge") +
  scale_fill_manual(values = c("Not Living in a Multigenerational Home" = "black","Living in a Multigenerational Home" = "#CFB991")) +
  labs(x = "Timepoint",y = "Frequency", fill = "Groups"
  ) +
  theme_minimal()

plot_nominal(
  data = subset_demo.multigen,
  var = multigen_label,
  title = "Multigenerational Household Distributions Across All Waves",
  xlab = "Multigenerational Household Status"
)

```

# Child Emotional/Behavioral Risk and Protective Factors

Child internalizing and externalizing behaviors were measured via
checklist measures.

## Dataset(s) & Variables

**mh_y_bpm**: Youth-reported externalizing and internalizing symptoms

Externalizing: Taken from 7 questions on the Brief Problem Monitor (BPM)
which were measured on a 3-point Likert scale with 0=not true, 2 =
somewhat true, 2= very true, 777 = decline to answer, 999 = don't know.

\*mh_y_bpm\_\_ext_001\* (formerly bpm_2_y): I argue a lot

\*mh_y_bpm\_\_ext_002\* (formerly bpm_6_y): I destroy things belonging
to others

\*mh_y_bpm\_\_ext_003\* (formerly bpm_7_y): I disobey my parents

\*mh_y_bpm\_\_ext_004\* (formerly bpm_8_y): I disobey at school. Please
select " Not True " if you are not in school at the time of this
assessment.

\*mh_y_bpm\_\_ext_005\* (formerly bpm_15_y): I am stubborn

\*mh_y_bpm\_\_ext_006\* (formerly bpm_16_y): I have a hot temper

\*mh_y_bpm\_\_ext_007\* (formerly bpm_17_y): I threaten to hurt people

\*mh_y_bpm\_\_ext_sum\* (formerly bpm_y_scr_external_r): Sum
[Validation: No more than 0 missing or declined]

\*mh_y_bpm\_\_ext_tscore\* (formerly bpm_y_scr_external_t): T-score
[Validation: No more than 0 missing or declined]

\*mh_y_bpm\_\_ext_nm\* (formerly bpm_y_scr_external_nm): Number missing
(externalizing)

Internalizing: Taken from 6 questions on the Brief Problem Monitor (BPM)
which were measured on a 3-point Likert scale with 0 = not true, 1 =
somewhat true, 2 = very true, 777 = decline to answer, 999 = don't know.

\*mh_y_bpm\_\_int_001\*: I feel worthless or inferior.

\*mh_y_bpm\_\_int_002\*: I am too fearful or anxious.

\*mh_y_bpm\_\_int_003\*: I feel too guilty.

\*mh_y_bpm\_\_int_004\*: I am self-conscious or easily embarrassed.

\*mh_y_bpm\_\_int_005\*: I am unhappy, sad, or depressed.

\*mh_y_bpm\_\_int_006\*: I worry a lot.

\*mh_y_bpm\_\_int_sum\*: Sum [Validation: No more than 0 missing or
declined]

\*mh_y_bpm\_\_int_tscore\*: T-score [Validation: No more than 0 missing
or declined]

\*mh_y_bpm\_\_int_nm\*: Number missing (internalizing)

**mh_p_cbcl**: Parent-reported externalizing and internalizing symptoms

Parent-reported externalizing symptoms were taken from the Child
Behavior Checklist (CBCL), which includes rule-breaking and aggressive
subscales. Parents indicted responses on a 3-point Likert scale with
0=not true and 2= very true. In order to match CBCL with BPM, we summed
the rule-breaking and aggressive subscales of the CBCL to create an
externalizing composite that excludes attention problems.

\*mh_p_cbcl\_\_synd\_\_ext_sum\*: Child Behavior Checklist [Parent]
(Syndrome Scale - Externalizing): Sum of the rule-breaking behavior and
aggressive behavior syndrome scales, excluding attention problems.

\*mh_p_cbcl\_\_synd\_\_ext_nm\*: Child Behavior Checklist [Parent]
(Syndrome Scale - Externalizing): Number missing.

Parent-reported internalizing symptoms were taken from the Child
Behavior Checklist (CBCL), which includes anxious/depressed,
withdrawn/depressed, and somatic complaints syndrome scales. Parents
indicated responses on a 3-point Likert scale with 0 = not true, 1 =
somewhat or sometimes true, and 2 = very true.

\*mh_p_cbcl\_\_synd\_\_int_sum\*: Child Behavior Checklist [Parent]
(Syndrome Scale - Internalizing): Sum of the anxious/depressed,
withdrawn/depressed, and somatic complaints syndrome scales.

\*mh_p_cbcl\_\_synd\_\_int_nm\*: Child Behavior Checklist [Parent]
(Syndrome Scale - Internalizing): Number missing.

Parent-reported attention problems were taken from the Child Behavior
Checklist (CBCL) attention problems syndrome scale. Parents indicated
responses on a 3-point Likert scale with 0 = not true, 1 = somewhat or
sometimes true, and 2 = very true.

\*mh_p_cbcl\_\_synd\_\_attn_sum\*: Child Behavior Checklist [Parent]
(Syndrome Scale - Attention problems): Sum [Validation: No more than 0
missing or declined]

\*mh_p_cbcl\_\_synd\_\_attn_nm\*: Child Behavior Checklist [Parent]
(Syndrome Scale - Attention problems): Number missing

**mh_y_pai**:Youth-reported positive affect

Youth-reported positive affect was assessed using the NIH Toolbox
Positive Affect Items [Youth]. Items capture the extent to which youth
report positive emotional experiences. The sum score was retained as an
indicator of youth positive affect.

*mh_y_pai_sum*: NIH Toolbox - Positive Affect Items [Youth]: Sum
[Validation: None missing or declined]

*mh_y_pai_nm*: NIH Toolbox - Positive Affect Items [Youth]: Number
missing


```{r , message=FALSE}
mental_health_codebook <- tibble::tribble(
  ~construct, ~indicator, ~description, ~waves_measured,

  "Mental health", "<i>mh_y_bpm__ext_sum</i>",
  "Youth-reported Brief Problem Monitor (BPM) externalizing symptoms sum score. Based on 7 items assessing arguing, rule-breaking, aggression, stubbornness, temper, and threats toward others.",
  "Waves 1-6",

  "Mental health", "<i>mh_y_bpm__ext_tscore</i>",
  "Youth-reported Brief Problem Monitor (BPM) externalizing symptoms T-score.",
  "Waves 1-6",

  "Mental health", "<i>mh_y_bpm__int_sum</i>",
  "Youth-reported Brief Problem Monitor (BPM) internalizing symptoms sum score. Based on 6 items assessing anxiety, guilt, embarrassment, sadness/depression, worthlessness, and worry.",
  "Waves 1-6",

  "Mental health", "<i>mh_y_bpm__int_tscore</i>",
  "Youth-reported Brief Problem Monitor (BPM) internalizing symptoms T-score.",
  "Waves 1-6",

  "Mental health", "<i>mh_p_cbcl__synd__ext_sum</i>",
  "Parent-reported Child Behavior Checklist (CBCL) externalizing syndrome scale sum score. Composite of the rule-breaking behavior and aggressive behavior syndrome scales, excluding attention problems.",
  "All waves (0-7)",

  "Mental health", "<i>mh_p_cbcl__synd__int_sum</i>",
  "Parent-reported Child Behavior Checklist (CBCL) internalizing syndrome scale sum score. Composite of the anxious/depressed, withdrawn/depressed, and somatic complaints syndrome scales.",
  "All waves (0-7)",

  "Mental health", "<i>mh_p_cbcl__synd__attn_sum</i>",
  "Parent-reported Child Behavior Checklist (CBCL) attention problems syndrome scale sum score.",
  "All waves (0-7)",

  "Positive affect", "<i>mh_y_pai_sum</i>",
  "Youth-reported NIH Toolbox Positive Affect Items sum score. Higher values indicate greater positive emotional experiences and positive affect.",
  "Waves 1, 3, and 5"
)

mental_health_codebook %>%
  dplyr::arrange(
    factor(
      construct,
      levels = c("Mental health", "Positive affect")
    ),
    indicator
  ) %>%
  kableExtra::kbl(
    format = "html",
    col.names = c("Construct", "Variable Name", "Variable Description", "Waves Measured"),
    escape = FALSE
  ) %>%
  kableExtra::kable_styling(
    bootstrap_options = c("hover", "condensed"),
    full_width = TRUE
  ) %>%
  kableExtra::collapse_rows(
    columns = 1,
    valign = "top"
  )

```


## Data Cleaning & Variable Creation

```{r extdata, message=FALSE}

#Subset brief problem monitor dataset to variables of interest
s.bpm <- c("participant_id", "session_id", "mh_y_bpm__ext_sum", "mh_y_bpm__int_sum")
bpm_subset <- bpm[s.bpm]

#Subset CBCL dataset to variables of interest
s.cbcl <- c("participant_id", "session_id", "mh_p_cbcl__synd__ext_sum", "mh_p_cbcl__synd__int_sum", "mh_p_cbcl__synd__attn_sum")
cbcl_subset <- cbcl[s.cbcl]

# Subset PAI dataset to variables of interest
s.pai <- c("participant_id", "session_id","mh_y_pai_sum")
pai_subset <- pai[s.pai]

#merge
mh_checklists <- merge(cbcl_subset, bpm_subset, by=c("participant_id", "session_id"), all = TRUE)
mh_checklists <- merge(mh_checklists,pai_subset,by = c("participant_id", "session_id"),all = TRUE)

#Drop assessments that are not annual assessments (indicated by M at the end of the session_id)
mh <- mh_checklists %>% 
  filter(session_id == "ses-00A" | session_id == "ses-01A" | session_id == "ses-02A" | session_id == "ses-03A" | session_id == "ses-04A" | session_id == "ses-05A" | session_id == "ses-06A")

mh = mh %>% 
  mutate(timepoint = str_extract(session_id, "\\d+") %>% as.integer())

```

## Descriptive Statistics

```{r, echo=FALSE, warning=FALSE, message=FALSE}

#Make a wide dataset
mhwide = mh %>%
  pivot_wider(
    id_cols = "participant_id", 
    names_from = "timepoint", 
    values_from = c("mh_y_bpm__ext_sum", "mh_y_bpm__int_sum","mh_p_cbcl__synd__ext_sum", "mh_p_cbcl__synd__int_sum", "mh_p_cbcl__synd__attn_sum","mh_y_pai_sum"))

kbl(psych::describe(mhwide), caption = "Child and Parent Reported Internalizing and Externalizing Summary Statistics") %>% 
  kable_minimal(c("striped", "hover","condensed","responsive")) %>%
  scroll_box(width = "100%")
```

## Viz: Distributions across all waves

```{r, echo=FALSE, warning=FALSE, message=FALSE}

plot_distribution(
     data = mh,
     var = mh_y_bpm__ext_sum,
     title = "Child Reported Externalizing",
     xlab = "Externalizing Problems",
     annotate_x = 5
   )

plot_distribution(
     data = mh,
     var = mh_y_bpm__int_sum,
     title = "Child Reported Internalizing",
     xlab = "Internalizing Problems",
     annotate_x = 5
   )

plot_distribution(
  data = mh,
  var = mh_p_cbcl__synd__ext_sum,
  title = "Parent Reported Externalizing",
  xlab = "Externalizing Problems",
  annotate_x = 10
)

plot_distribution(
  data = mh,
  var = mh_p_cbcl__synd__int_sum,
  title = "Parent Reported Internalizing",
  xlab = "Internalizing Problems",
  annotate_x = 15
)

plot_distribution(
  data = mh,
  var = mh_p_cbcl__synd__attn_sum,
  title = "Parent Reported Attention Problems",
  xlab = "Attention Problems",
  annotate_x = 8
)

plot_distribution(
  data = mh,
  var = mh_y_pai_sum,
  title = "Child Reported Positive Affect",
  xlab = "Positive Affect",
  annotate_x = 10
)

```

## Viz: Distributions by wave

```{r, echo=FALSE, warning=FALSE, message=FALSE}

plot_distribution_by_time(
  data = mh,
  var = mh_y_bpm__ext_sum,
  time_var = session_id,
  title = "Child Reported Externalizing",
  xlab = "Externalizing Problems",
  annotate_x = 5
)

plot_distribution_by_time(
  data = mh,
  var = mh_y_bpm__int_sum,
  time_var = session_id,
  title = "Child Reported Internalizing by Wave",
  xlab = "Internalizing Problems",
  annotate_x = 5
)

plot_distribution_by_time(
  data = mh,
  var = mh_p_cbcl__synd__ext_sum,
  time_var = session_id,
  title = "Parent Reported Externalizing by Wave",
  xlab = "Externalizing Problems",
  annotate_x = 10
)

plot_distribution_by_time(
  data = mh,
  var = mh_p_cbcl__synd__int_sum,
  time_var = session_id,
  title = "Parent Reported Internalizing by Wave",
  xlab = "Internalizing Problems",
  annotate_x = 10
)

plot_distribution_by_time(
  data = mh,
  var = mh_p_cbcl__synd__attn_sum,
  time_var = session_id,
  title = "Parent Reported Attention Problems by Wave",
  xlab = "Attention Problems",
  annotate_x = 5
)

plot_distribution_by_time(
  data = mh,
  var = mh_y_pai_sum,
  time_var = session_id,
  title = "Child Reported Positive Affect by Wave",
  xlab = "Positive Affect",
  annotate_x = 20
)
```

## Viz: Longitudinal Plots

```{r, echo=FALSE, warning=FALSE, message=FALSE}

plot_longitudinal_subset(
  data = mh,
  var = mh_y_bpm__ext_sum,
  time_var = timepoint,
  id_var = participant_id,
  sample_n = 100,
  seed = 2025,  
  title = "Child Reported Externalizing over Time",
  xlab = "Time",
  ylab = "Externalizing Problems"
)

plot_longitudinal_subset(
  data = mh,
  var = mh_y_bpm__int_sum,
  time_var = timepoint,
  id_var = participant_id,
  sample_n = 100,
  seed = 2025,
  title = "Child Reported Internalizing over Time",
  xlab = "Time",
  ylab = "Internalizing Problems"
)

plot_longitudinal_subset(
  data = mh,
  var = mh_p_cbcl__synd__ext_sum,
  time_var = timepoint,
  id_var = participant_id,
  sample_n = 100,
  seed = 2025,
  title = "Parent Reported Externalizing over Time",
  xlab = "Time",
  ylab = "Externalizing Problems"
)

plot_longitudinal_subset(
  data = mh,
  var = mh_p_cbcl__synd__int_sum,
  time_var = timepoint,
  id_var = participant_id,
  sample_n = 100,
  seed = 2025,
  title = "Parent Reported Internalizing over Time",
  xlab = "Time",
  ylab = "Internalizing Problems"
)

plot_longitudinal_subset(
  data = mh,
  var = mh_p_cbcl__synd__attn_sum,
  time_var = timepoint,
  id_var = participant_id,
  sample_n = 100,
  seed = 2025,
  title = "Parent Reported Attention Problems over Time",
  xlab = "Time",
  ylab = "Attention Problems"
)

plot_longitudinal_subset(
  data = mh,
  var = mh_y_pai_sum,
  time_var = timepoint,
  id_var = participant_id,
  sample_n = 100,
  seed = 2025,
  title = "Child Reported Positive Affect over Time",
  xlab = "Time",
  ylab = "Positive Affect"
)
```

# Substance Use

We constructed uptake continuum measures for 1) alcohol, 2) tobacco, and
3) marijuana.

Continuum scoring depended upon items available at each assessment
(e.g., follow-ups ask about the last 12 months, baseline is lifetime).
Categories were collapsed or expanded depending on rates of endorsement
but generally follow patterns of milestone attainment.

## Datasets & Variables

Alcohol

**Su_y_sui**

\*su_y_sui\_\_hrd\_\_alc_001\* (formerly tlfb_alc & tlfb_alc_l):Have you
heard of----alcohol, such as beer, wine, or liquor? 0 = No; 1 = Yes
(NOTE: full data only available for ses-00A, 409 responses available for
ses-01A, 140 responses available for ses-02A)
\*su_y_sui\_\_use\_\_alc_001\* & \*su_y_sui\_\_use\_\_alc_001\_\_l\*
(formerly tlfb_alc_use & tlfb_alc_use_l): Have you ever used: A full
drink of beer, wine or liquor (rum, vodka, gin, whiskey) 0 = No; 1 = Yes

**su_y_lowuse**

\*su_y_lowuse\_\_isip_002\* & \*su_y_lowuse\_\_isip_002\_\_l\*: A sip of
alcohol such as beer, wine or liquor (rum, vodka, gin, whiskey).0 = No;
1 = Yes

**su_y_itu**

\*su_y_itu\_\_alc_001\* (formerly path_alc_youth2 & path_alc_youth2_l):
Have you ever been curious about drinking alcohol? 1=Very curious; 2
Somewhat curious; 3 = A little curious; 4 = Not at all curious; 999=
Don’t know; 777 = Refused to answer \*su_y_itu\_\_alc_002\* (formerly
path_alc_youth5 & path_alc_youth5_l): Do you think you will try alcohol
soon? 1 = Definitely yes; 2 = Probably yes; 3 = Probably not; 4=
Definitely not; 999= Don’t know; 777 = Refused to answer
\*su_y_itu\_\_alc_003\* (formerly path_alc_youth8 & path_alc_youth8_l):
If one of your best friends were to offer you alcohol, would you try it?
1 = definitely yes; 2 = probably yes; 3 = Probably Not; 4 = definitely
not; 999= Don’t know; 777 = Refused to answer

**su_y_tlfb**

\*su_y_tlfb\_\_alc\_\_binge_ud\* (formerly tlfb_cal_scr_num_bde):
Alcohol - Binge episodes: Number of use days \*su_y_tlfb\_\_alc_ud\*
(formerly tlfb_cal_scr_alc_ud): Alcohol: Number of use days
\*su_y_tlfb\_\_alc_mean\* (formerly tlfb_cal_scr_alc_avg): Alcohol:
Average dose

**su_y_alcsre**

\*su_y_alcsre\_\_6mo_count\* (formerly subj_resp_alc_total_6mo &
subj_resp_alc_total_6mo_i): Number of subjective experiences from
drinking alcohol during in the past 6 months
\*su_y_alcsre\_\_hvy_count\* (formerly subj_resp_alc_total_heavy &
subj_resp_alc_total_heavy_i): Number of subjective experiences from
drinking alcohol during their heaviest drinking period.

Tobacco

**Su_y_sui**

\*su_y_sui\_\_hrd\_\_nic_001\* (formerly tlfb_tob & tlfb_tob_l): Have
you heard of tobacco products, such as cigarettes, smokeless tobacco,
cigars, hookah, electronic or e-cigarettes, or Juul? 0 = No; 1 = Yes
(NOTE: full data only available for ses-00A, 712 reponses for ses-01A,
266 responses for ses-02A) \*su_y_sui\_\_use\_\_nic\_\_cig_001\* &
\*su_y_sui\_\_use\_\_nic\_\_cig_001\_\_l\* (formerly tlfb_cig_use &
tlfb_cig_use_l): Tobacco cigarette - more than just a puff.0 = No; 1 =
Yes

**su_y_lowuse**

\*su_y_lowuse\_\_nic_001\* & \*su_y_lowuse\_\_nic_001\_\_l\* (formerly
first_nicotine_1 & first_nicotine_1_l): Have you had a puff of a
cigarette, electronic nicotine or vaping products, such as e-cigarettes,
vape pens, JUULs, hookah, cigar or pipe. 0 = No; 1 = Yes

**su_y_itu**

\*su_y_itu\_\_nic_001\* (formerly path_alc_youth1 & path_alc_youth1_l):
Have you ever been curious about using a tobacco product such as
cigarettes, e-cigarettes, hookah, or cigars? 1 = Very curious; 2 =
Somewhat curious; 3 = A little curious; 4 = Not at all curious; 999=
Don’t know; 777 = Refused to answer \*su_y_itu\_\_nic_002\* (formerly
path_alc_youth4 & path_alc_youth4_l): Do you think you will try a
tobacco product soon? 1 = Definitely yes; 2 = Probably yes; 3 = Probably
not; 4 = Definitely not; 999= Don’t know; 777 = Refused to answer
\*su_y_itu\_\_nic_003\* (formerly path_alc_youth7 & path_alc_youth7_l):
If one of your best friends were to offer you a tobacco product, would
you try it? 1 = Definitely yes; 2 = Probably yes; 3 = Probably not; 4 =
Definitely not; 999= Don’t know; 777 = Refused to answer

**su_y_tlfb**

\*su_y_tlfb\_\_nic\_\_cig_mean\* (formerly tlfb_cal_scr_cig_avg):
Nicotine - Cigarettes: Average dose \*su_y_tlfb\_\_nic\_\_cig_ud\*
(formerly tlfb_cal_scr_cig_ud): Nicotine - Cigarettes: Number of use
days

**su_y_nicsre**

\*su_y_nicsre\_\_cig_sum\* (formerly res_tobacco_total_cig &
res_tobacco_total_cig_i): Total intensity of positive and negative
effects of first cigarette use.

Marijuana

**Su_y_sui**

\*su_y_sui\_\_hrd\_\_mj_001\* (formerly tlfb_mj & tlfb_mj_l): Have you
ever heard of marijuana, weed, pot, blunts, dabs, marijuana drinks or
food with marijuana? 0 = No; 1 = Yes (NOTE: full data only available for
ses-00A, 4888 response available for ses-01A, 2375 responses available
for ses-02A) \*su_y_sui\_\_use\_\_mj\_\_puff_001\* &
\*su_y_sui\_\_use\_\_mj\_\_puff_001\_\_l\* (formerly tlfb_mj_puff &
tlfb_mj_puff_l): Have you ever used: A puff or eaten any marijuana, also
called pot, grass, weed or ganja? 0 = No; 1 = Yes
\*su_y_sui\_\_use\_\_mj\_\_smoke_001\* &
\*su_y_sui\_\_use\_\_mj\_\_smoke_001\_\_l\* (formerly tlfb_mj_use &
tlfb_mj_use_l): Have you ever used: Smoked marijuana, also called pot,
grass, weed, ganja - more than just a puff? 0 = No; 1 = Yes

**su_y_tlfb**

\*su_y_tlfb\_\_mj\_\_smoke_mean\* (formerly tlfb_cal_scr_smj_avg):
Average smoked MJ on use days \*su_y_tlfb\_\_mj\_\_smoke_ud\* (formerly
tlfb_cal_scr_smj_ud): Marijuana - Smoked: Number of use days

**su_y_itu**

\*su_y_itu\_\_mj_001\* (formerly path_alc_youth3 & path_alc_youth3_l):
Have you ever been curious about trying marijuana? 1 = Very curious; 2 =
Somewhat curious; 3 = A little curious; 4 = Not at all curious; 999=
Don’t know; 777 = Refused to answer \*su_y_itu\_\_mj_002\* (formerly
path_alc_youth6 & path_alc_youth6_l): Do you think you will try
marijuana soon? 1 = Definitely yes; 2 = Probably yes; 3 = Probably not;
4 = Definitely not; 999= Don’t know; 777 = Refused to answer
\*su_y_itu\_\_mj_003\* (formerly path_alc_youth9 & path_alc_youth9_l):
If one of your best friends were to offer you marijuana, would you try
it? 1 = Definitely yes; 2 = Probably yes; 3 = Probably not; 4 =
Definitely not; 999= Don’t know; 777 = Refused to answer

**su_y_mjsre**

*su_y_mjsre_sum* (formerly sub_res_mj_total & sub_res_mj_total_i): total
number and severity of subjective responses for first marijuana use
episode.

## Data Cleaning

```{r sucleaning, message=FALSE}


#Subset sui dataset to variables of interest
s.sui <- c("participant_id", "session_id", "su_y_sui__hrd__alc_001", "su_y_sui__use__alc_001", "su_y_sui__use__alc_001__l", "su_y_sui__hrd__nic_001", "su_y_sui__use__nic__cig_001", "su_y_sui__use__nic__cig_001__l", "su_y_sui__hrd__mj_001", "su_y_sui__use__mj__puff_001", "su_y_sui__use__mj__puff_001__l", "su_y_sui__use__mj__smoke_001", "su_y_sui__use__mj__smoke_001__l")
sui_subset <- sui[s.sui]

#Subset lowuse dataset to variables of interest
s.lowuse <- c("participant_id", "session_id", "su_y_lowuse__isip_002", "su_y_lowuse__isip_002__l", "su_y_lowuse__nic_001", "su_y_lowuse__nic_001__l")
lowuse_subset <- lowuse[s.lowuse]

#Subset intention to use (itu) dataset to variables of interest
s.itu <- c("participant_id", "session_id", "su_y_itu__alc_001", "su_y_itu__alc_002", "su_y_itu__alc_003", "su_y_itu__nic_001", "su_y_itu__nic_002", "su_y_itu__nic_003", "su_y_itu__mj_001", "su_y_itu__mj_002", "su_y_itu__mj_003")
itu_subset <- itu[s.itu]

#Subset tlfb dataset to variables of interest
s.tlfb <- c("participant_id", "session_id", "su_y_tlfb__alc__binge_ud", "su_y_tlfb__alc_ud", "su_y_tlfb__alc_mean", "su_y_tlfb__nic__cig_mean", "su_y_tlfb__nic__cig_ud", "su_y_tlfb__mj__smoke_mean", "su_y_tlfb__mj__smoke_ud")
tlfb_subset <- tlfb[s.tlfb]

#Subset subjective experiences dataset to variables of interest
#alcohol
s.alcsre <- c("participant_id", "session_id", "su_y_alcsre__6mo_count", "su_y_alcsre__hvy_count")
alcsre_subset <- alcsre[s.alcsre]

#Tobacco
s.nicsre <- c("participant_id", "session_id", "su_y_nicsre__cig_sum")
nicsre_subset <- nicsre[s.nicsre]

#Marijuana
s.mjsre <- c("participant_id", "session_id", "su_y_mjsre_sum")
mjsre_subset <- mjsre[s.mjsre]

#combine all substance use variable datasets
substance_datavars <- list(sui_subset, lowuse_subset, itu_subset, tlfb_subset, alcsre_subset, nicsre_subset, mjsre_subset)
su_merged <- Reduce(function(x, y) merge(x, y, by = c("participant_id","session_id"), all = TRUE), substance_datavars)

#combine baseline and longitudinal variables
su_merged1 <- su_merged %>% 
  mutate(fulldrink = if_else(is.na(su_y_sui__use__alc_001), su_y_sui__use__alc_001__l, su_y_sui__use__alc_001)) %>% 
  mutate(sipalc = if_else(is.na(su_y_lowuse__isip_002), su_y_lowuse__isip_002__l, su_y_lowuse__isip_002)) %>% 
  mutate(puffnic = if_else(is.na(su_y_lowuse__nic_001), su_y_lowuse__nic_001__l, su_y_lowuse__nic_001)) %>% 
  mutate(fullnic = if_else(is.na(su_y_sui__use__nic__cig_001), su_y_sui__use__nic__cig_001__l, su_y_sui__use__nic__cig_001))%>% 
  mutate(puffmj = if_else(is.na(su_y_sui__use__mj__puff_001), su_y_sui__use__mj__puff_001__l, su_y_sui__use__mj__puff_001)) %>% 
  mutate(fullmj = if_else(is.na(su_y_sui__use__mj__smoke_001), su_y_sui__use__mj__smoke_001__l, su_y_sui__use__mj__smoke_001))

#Change path variables with 999 and 777 to NA
su_merged1 <- su_merged1 %>% 
mutate(curiousalc = if_else(su_y_itu__alc_001 %in% c(777,999), NA, su_y_itu__alc_001)) %>% 
mutate(tryalc = if_else(su_y_itu__alc_002 %in% c(777,999), NA, su_y_itu__alc_002)) %>% 
mutate(offeralc = if_else(su_y_itu__alc_003 %in% c(777,999), NA, su_y_itu__alc_003)) %>% 
mutate(curiouscig = if_else(su_y_itu__nic_001 %in% c(777,999), NA, su_y_itu__nic_001)) %>% 
mutate(trycig = if_else(su_y_itu__nic_002 %in% c(777,999), NA, su_y_itu__nic_002)) %>% 
mutate(offercig = if_else(su_y_itu__nic_003 %in% c(777,999), NA, su_y_itu__nic_003)) %>% 
mutate(curiousmj = if_else(su_y_itu__mj_001 %in% c(777,999), NA, su_y_itu__mj_001)) %>% 
mutate(trymj = if_else(su_y_itu__mj_002 %in% c(777,999), NA, su_y_itu__mj_002)) %>% 
mutate(offermj = if_else(su_y_itu__mj_003 %in% c(777,999), NA, su_y_itu__mj_003))

```

## Variable Creation

**Definitions of Uptake Continuums**

0 = never heard of substance Alcohol: su_y_sui\_\_hrd\_\_alc_001 (1=1)
Tobacco: su_y_sui\_\_hrd\_\_nic_001 (1=1) Marijuana:
su_y_sui\_\_hrd\_\_mj_001 (1=1)

1 = Curious about substance use Alcohol: curiousalc (\<3=1) OR tryalc
(\<3=1) Tobacco: curiouscig (\<3=1) OR trycig (\<3=1) Marijuana:
curiousmj (\<3=1) OR trymj (\<3=1)

2 = Susceptible substance user Alcohol: offeralc (\<3=1) Tobacco:
offercig (\<3=1) Marijuana: offermj (\<3=1)

3 = Initiated (sip or puff) Alcohol: sipalc (1=1) Tobacco: puffnic (1=1)
Marijuana: puffmj (1=1)

4 = Whole Portion (full drink, whole cigarette, whole joint) Alcohol:
fulldrink (1=1) Tobacco: fullnic (1=1) Marijuana: fullmj (1=1)

5 = regular use Alcohol: su_y_tlfb\_\_alc_ud (\>=26; drank once per week
over 6 months) Tobacco: su_y_tlfb\_\_nic\_\_cig_ud (\>=26; smoked once
per week over 6 months) Marijuana: su_y_tlfb\_\_mj\_\_smoke_ud (\>=26;
smoked once per week over 6 months)

6 = Experience Effects of Substance Alcohol:su_y_alcsre\_\_6mo_count
(\>0=1) OR su_y_alcsre\_\_hvy_count (\>0=1) Tobacco:
su_y_nicsre\_\_cig_sum (\>0=1) Marijuana: su_y_mjsre_sum (\>0=1)

7 = Heavy Use Alcohol: su_y_tlfb\_\_alc\_\_binge_ud (\>0=1) OR
su_y_tlfb\_\_alc_mean (\>=4 drinks per occasion) OR su_y_tlfb\_\_alc_ud
((\>=52; drank twice per week over 6 months) Tobacco:
su_y_tlfb\_\_nic\_\_cig_mean (\>=25 cigarettes per occasion) OR
su_y_tlfb\_\_nic\_\_cig_ud (\>=52; smoked twice per week over 6 months)
Marijuana: su_y_tlfb\_\_mj\_\_smoke_mean (\>=1 joint per occassion) OR
su_y_tlfb\_\_mj\_\_smoke_ud (\>=52; smoked twice per week over 6 months)

```{r sucreation, message=FALSE}

#create separate alcohol use uptake variables 
su_merged2 <- su_merged1 %>%
mutate(AUC7 = ifelse((su_y_tlfb__alc__binge_ud > 0 | su_y_tlfb__alc_mean >=4 | su_y_tlfb__alc_ud >=52), 7, 0),
AUC6 = ifelse((su_y_alcsre__6mo_count >0 | su_y_alcsre__hvy_count >0), 6, 0),
AUC5 = ifelse((su_y_tlfb__alc_ud >=26), 5, 0),
AUC4 = ifelse(fulldrink == 1, 4, 0),
AUC3 = ifelse(sipalc == 1, 3, 0),
AUC2 = ifelse(offeralc %in% c(1,2), 2, 0),
AUC1 = ifelse((curiousalc %in% c(1,2) | tryalc %in% c(1,2)), 1, 0), 
AUC0 = ifelse(su_y_sui__hrd__alc_001 == 1, 0.01, 0))

#merge alcohol use uptake variables into one continuum taking the max
   su_merged2 <- su_merged2 %>% 
rowwise() %>% mutate(AUC = max(c(AUC0, AUC1, AUC2, AUC3, AUC4, AUC5, AUC6, AUC7), na.rm=T))
   
   table(su_merged2$AUC)

   
   #create separate cigarette use uptake variables 
su_merged2 <- su_merged2 %>%
mutate(TUC7 = ifelse((su_y_tlfb__nic__cig_mean >=25 | su_y_tlfb__nic__cig_ud >=52), 7, 0), #td : error -> avg
TUC6 = ifelse(su_y_nicsre__cig_sum >0, 6, 0),
TUC5 = ifelse((su_y_tlfb__nic__cig_ud >=26), 5, 0),
TUC4 = ifelse(fullnic == 1, 4, 0),
TUC3 = ifelse(puffnic == 1, 3, 0),
TUC2 = ifelse(offercig %in% c(1,2), 2, 0),
TUC1 = ifelse((curiouscig %in% c(1,2) | trycig %in% c(1,2)), 1, 0), 
TUC0 = ifelse(su_y_sui__hrd__nic_001 == 1, 0.01, 0))

#merge cigarette use uptake variables into one continuum taking the max
   su_merged2 <- su_merged2 %>% 
rowwise() %>% mutate(TUC= if_else(all(is.na(TUC0), is.na(TUC1), is.na(TUC2), is.na(TUC3), is.na(TUC4), is.na(TUC5), is.na(TUC6), is.na(TUC7)), NA, max(TUC0, TUC1, TUC2, TUC3, TUC4, TUC5, TUC6, TUC7, na.rm=T)))
   
   table(su_merged2$TUC)
   
   #create separate marijuana use uptake variables 
su_merged2 <- su_merged2 %>%
mutate(MUC7 = ifelse((su_y_tlfb__mj__smoke_mean >=1 | su_y_tlfb__mj__smoke_ud >=52), 7, 0),
MUC6 = ifelse(su_y_mjsre_sum >0, 6, 0),
MUC5 = ifelse((su_y_tlfb__mj__smoke_ud >=26), 5, 0),
MUC4 = ifelse(fullmj == 1, 4, 0),
MUC3 = ifelse(puffmj == 1, 3, 0),
MUC2 = ifelse(offermj %in% c(1,2), 2, 0),
MUC1 = ifelse((curiousmj %in% c(1,2) | trymj %in% c(1,2)), 1, 0), 
MUC0 = ifelse(su_y_sui__hrd__mj_001 == 1, 0.01, 0))

#merge marijuana use uptake variables into one continuum taking the max
   su_merged2<- su_merged2 %>% 
rowwise() %>% mutate(MUC= if_else(all(is.na(MUC0), is.na(MUC1), is.na(MUC2), is.na(MUC3), is.na(MUC4), is.na(MUC5), is.na(MUC6), is.na(MUC7)), NA, max(MUC0, MUC1, MUC2, MUC3, MUC4, MUC5, MUC6, MUC7, na.rm=T)))
   
   table(su_merged2$MUC)

# Create overall substance use involvement indicator and initiation variables:

su_merged2 <- su_merged2 %>%
  # recode any 0.01 to 0
  mutate(
    AUC = ifelse(AUC == 0.01, 0, AUC),
    TUC = ifelse(TUC == 0.01, 0, TUC),
    MUC = ifelse(MUC == 0.01, 0, MUC)
  ) %>%
  # create *_init variables (1 if >=1, 0 if 0)
  mutate(
    AU_init = ifelse(AUC >= 3, 1, 0),
    TU_init = ifelse(TUC >= 3, 1, 0),
    MU_init = ifelse(MUC >= 3, 1, 0),
    SU_init = ifelse(AU_init == 1 | TU_init == 1 | MU_init == 1, 1, 0)
  ) %>%
  # create overall substance use involvement
  mutate(
    SUC = rowSums(across(c(AUC, TUC, MUC)), na.rm = FALSE)
  )

#pairdown to just outcome variables (i.e., AUC, TUC, MUC)
s.su1 <- c("participant_id", "session_id", "AUC", "TUC", "MUC", "SUC", "AU_init","TU_init","MU_init","SU_init","AUC0", "AUC1", "AUC2", "AUC3", "AUC4", "AUC5", "AUC6", "AUC7")
su <- su_merged2[s.su1]
   
#Note: Final dataframe name is su

```

## Descriptive Statistics

```{r, echo=FALSE, warning=FALSE, message=FALSE}

#Make a wide dataset
suwide = su %>%
  pivot_wider(
    id_cols = "participant_id", 
    names_from = "session_id", 
    values_from = c("AUC", "TUC", "MUC", "SUC", "AU_init","TU_init","MU_init","SU_init"))

kbl(psych::describe(suwide ), caption = "Substance Use Continuum Summary Statistics") %>% 
  kable_minimal(c("striped", "hover","condensed","responsive")) %>%
  scroll_box(width = "100%")

```

## Viz: Distributions across all waves

```{r, echo=FALSE, warning=FALSE, message=FALSE}

plot_distribution(
  data = su,
  var = AUC,
  title = "Alcohol Uptake Continuum",
  xlab = "Alcohol Uptake Continuum",
  annotate_x = 2.5
)

plot_distribution(
  data = su,
  var = TUC,
  title = "Tobacco Uptake Continuum",
  xlab = "Tobacco Uptake Continuum",
  annotate_x = 2.5
)

plot_distribution(
  data = su,
  var = MUC,
  title = "Marijuana Uptake Continuum",
  xlab = "Marijuana Uptake Continuum",
  annotate_x = 2
)

plot_distribution(
  data = su,
  var = SUC,
  title = "Overall Substance Use Involvement",
  xlab = "Substance Use Involvement",
  annotate_x = 5
)

plot_distribution(
  data = su,
  var = AU_init,
  title = "Alcohol Use Initiation",
  xlab = "Alcohol Use Initiation",
  annotate_x = 0.75
)

plot_distribution(
  data = su,
  var = TU_init,
  title = "Tobacco Use Initiation",
  xlab = "Tobacco Use Initiation",
  annotate_x = 0.26
)

plot_distribution(
  data = su,
  var = MU_init,
  title = "Marijuana Use Initiation",
  xlab = "Marijuana Use Initiation",
  annotate_x = 0.5
)

plot_distribution(
  data = su,
  var = SU_init,
  title = "Any Substance Use Initiation",
  xlab = "Any Substance Use Initiation",
  annotate_x = 0.9
)

```

## Viz: Distributions by wave

```{r, echo=FALSE, warning=FALSE, message=FALSE}

plot_distribution_by_time(
  data = su,
  var = AUC,
  time_var = session_id,
  title = "Alcohol Uptake Continuum by Wave",
  xlab = "Alcohol Uptake Continuum",
  annotate_x = 2
)

plot_distribution_by_time(
  data = su,
  var = TUC,
  time_var = session_id,
  title = "Tobacco Uptake Continuum by Wave",
  xlab = "Tobacco Uptake Continuum",
  annotate_x = 2
)

plot_distribution_by_time(
  data = su,
  var = MUC,
  time_var = session_id,
  title = "Marijuana Uptake Continuum by Wave",
  xlab = "Marijuana Uptake Continuum",
  annotate_x = 2
)

plot_distribution_by_time(
  data = su,
  var = SUC,
  time_var = session_id,
  title = "Overall Substance Use Involvement by Wave",
  xlab = "Substance Use Involvement",
  annotate_x = 2
)

plot_distribution_by_time(
  data = su,
  var = AU_init,
  time_var = session_id,
  title = "Alcohol Use Initiation by Wave",
  xlab = "Alcohol Use Initiation",
  annotate_x = 0.5
)

plot_distribution_by_time(
  data = su,
  var = TU_init,
  time_var = session_id,
  title = "Tobacco Use Initiation by Wave",
  xlab = "Tobacco Use Initiation",
  annotate_x = 0.5
)

plot_distribution_by_time(
  data = su,
  var = MU_init,
  time_var = session_id,
  title = "Marijuana Use Initiation by Wave",
  xlab = "Marijuana Use Initiation",
  annotate_x = 0.5
)

plot_distribution_by_time(
  data = su,
  var = SU_init,
  time_var = session_id,
  title = "Any Substance Use Initiation by Wave",
  xlab = "Any Substance Use Initiation",
  annotate_x = 0.5
)
```

## Viz: Longitudinal Plots

```{r, echo=FALSE, warning=FALSE, message=FALSE}


su <- su %>% 
  mutate(timepoint = str_extract(session_id, "\\d+") %>% as.integer())

plot_longitudinal_subset(
  data = su,
  var = AUC,
  time_var = timepoint,
  id_var = participant_id,
  sample_n = 100,
  seed = 2025,
  title = "Alcohol Uptake Continuum over Time",
  xlab = "Time",
  ylab = "Alcohol Uptake Continuum"
)

plot_longitudinal_subset(
  data = su,
  var = TUC,
  time_var = timepoint,
  id_var = participant_id,
  sample_n = 100,
  seed = 2025,
  title = "Tobacco Uptake Continuum over Time",
  xlab = "Time",
  ylab = "Tobacco Uptake Continuum"
)

plot_longitudinal_subset(
  data = su,
  var = MUC,
  time_var = timepoint,
  id_var = participant_id,
  sample_n = 100,
  seed = 2025,
  title = "Marijuana Uptake Continuum over Time",
  xlab = "Time",
  ylab = "Marijuana Uptake Continuum"
)

plot_longitudinal_subset(
  data = su,
  var = SUC,
  time_var = timepoint,
  id_var = participant_id,
  sample_n = 100,
  seed = 2025,
  title = "Overall Substance Use Involvement over Time",
  xlab = "Time",
  ylab = "Substance Use Involvement"
)

plot_longitudinal_subset(
  data = su,
  var = AU_init,
  time_var = timepoint,
  id_var = participant_id,
  sample_n = 100,
  seed = 2025,
  title = "Alcohol Use Initiation over Time",
  xlab = "Time",
  ylab = "Alcohol Use Initiation"
)

plot_longitudinal_subset(
  data = su,
  var = TU_init,
  time_var = timepoint,
  id_var = participant_id,
  sample_n = 100,
  seed = 2025,
  title = "Tobacco Use Initiation over Time",
  xlab = "Time",
  ylab = "Tobacco Use Initiation"
)

plot_longitudinal_subset(
  data = su,
  var = MU_init,
  time_var = timepoint,
  id_var = participant_id,
  sample_n = 100,
  seed = 2025,
  title = "Marijuana Use Initiation over Time",
  xlab = "Time",
  ylab = "Marijuana Use Initiation"
)

plot_longitudinal_subset(
  data = su,
  var = SU_init,
  time_var = timepoint,
  id_var = participant_id,
  sample_n = 100,
  seed = 2025,
  title = "Any Substance Use Initiation over Time",
  xlab = "Time",
  ylab = "Any Substance Use Initiation"
)
```

# Parent and Peer Variables

## Family Enviornment

### Datasets & Variables

**fc_y_fes**: Child-percived family cohesion and family conflict

\*fc_y_fes\_\_cohes_mean\*: Family Environment Scale [Youth] (Cohesion):
Mean [Validation: No more than 1 missing or declined]

\*fc_y_fes\_\_cohes_nm\*: Family Environment Scale [Youth] (Cohesion):
Number missing

\*fc_y_fes\_\_confl_mean\*: Family Environment Scale [Youth] (Conflict):
Mean [Validation: No more than 1 missing or declined]

\*fc_y_fes\_\_confl_nm\*: Family Environment Scale [Youth] (Conflict):
Number missing

## Peer Behavior Profile

### Datasets & Variables

**fc_y_pbp**: The Peer Behavior Profile [Youth] assesses youths’
perceptions of their friends’ behaviors and characteristics. Items were
measured on an ordinal response scale: 1 = None or almost none, 2 = A
few, 3 = Half, 4 = Most, 5 = All or almost all, and 999 = Don't know.
This section uses six peer behavior items. Three items reflect friends’
problem or risk behaviors, including skipping school, being suspended
from school, and shoplifting. Three items reflect positive or prosocial
peer characteristics, including being athletes, attending spiritual or
religious services, and being excellent students. These items can be
used to create negative peer and positive peer summary variables.

*fc_y_pbp_001*: How many of your friends have skipped school?

*fc_y_pbp_002*: How many of your friends have been suspended from
school?

*fc_y_pbp_003*: How many of your friends have shoplifted occasionally?

*fc_y_pbp_004*: How many of your friends are athletes?

*fc_y_pbp_005*: How many of your friends attend spiritual or religious
services once a month or more often?

*fc_y_pbp_006*: How many of your friends are excellent students (GPA 3.5
[B+] or higher)?

## Resistance to Peer Influence

### Datasets & Variables

**fc_y_rpi**: Child-reported Resistance to Peer Influence. The
Resistance to Peer Influence [Youth] scale assesses youths’ ability to
resist peer pressure and maintain autonomous decision-making in social
situations. Items present paired descriptions of youth behavior, such as
whether the youth tends to keep friends happy, fit in with the crowd,
change their mind because of peers, or go along with what friends do.
This section uses the mean summary score. Higher scores indicate greater
resistance to peer influence.

*fc_y_rpi_mean*: Resistance to Peer Influence [Youth]: Mean summary
score

## Parental Monitoring

### Datasets & Variables

**fc_y_pm**: Child-Perceived, General Monitoring / Knowledge

*fc_y_pm_001* (formerly parent_monitor_q1_y): "How often do your
parents/guardians know where you are". 1 = Never, 2 = Almost Never, 3 =
Sometimes, 4 = Often, 5 = Always or Almost Always, 777 = Decline to
answer *fc_y_pm_002* (formerly parent_monitor_q2_y): "How often do your
parents know who you are with when you are not at school and away from
home". 1 = Never, 2 = Almost Never, 3 = Sometimes, 4 = Often, 5 = Always
or Almost Always, 777 = Decline to answer *fc_y_pm_003* (formerly
parent_monitor_q3_y): "If you are at home when parents/guardians not,
how often do you know how to get in touch with them". 1 = Never, 2 =
Almost Never, 3 = Sometimes, 4 = Often, 5 = Always or Almost Always, 777
= Decline to answer *fc_y_pm_004* (formerly parent_monitor_q4_y): "How
often do you talk to your parent or guardian about your plans for the
coming day, such as your plans about what will happen at school or what
you are going to do with friends" 1 = Never, 2 = Almost Never, 3 =
Sometimes, 4 = Often, 5 = Always or Almost Always, 777 = Decline to
answer *fc_y_pm_005* (formerly parent_monitor_q5_y): "In an average
week, how many times do you and your parents/guardians eat dinner
together". 1 = Never, 2 = Almost Never, 3 = Sometimes, 4 = Often, 5 =
Always or Almost Always, 777 = Decline to answer

*fc_y_pm_mean* is the summary score.

## Parental Control

Was indexed primarily by rules about substance use endorsed by parents.
Items include rules about drinking, smoking, and marijuana use (reverse
coded). A second set of items assessed whether parents enforced
penalties for violating family rules about each substance. Higher scores
will reflect more, enforced rules.

### Datasets & Variables

**su_p_rule**

Alcohol

\*su_p_rule\_\_alc_001\* (formerly parent_rules_q1): Which statement
best describes the rules for your child's drinking in your house. 1 = My
child is not allowed to drink under any circumstances; 2 = My child is
allowed to drink at home but only on special occasions and under parent
supervision; 3 = My child can have a drink when he/she asks for one; 4 =
My child can drink at home whenever he/she wants to; 5 = I don't set
rules about my child's drinking; 6 = I haven't made rules yet about my
child drinking; 777 = Decline to answer

\*su_p_rule\_\_alc_001\_\_03\* (formerly parent_rules_q3): Are these
rules (regarding alcohol consumption) enforced? 0 = no; 1 = yes; 4 = I
have not had to enforce these rules, 777 = Decline to answer

Smoking

\*su_p_rule\_\_nic\_\_cig_001\* (formerly parent_rules_q4): What are the
family rules about smoking cigarettes ffor your child? 1 = Not allowed
to smoke under any circumstances; 2 = Not allowed to smoke in home - no
family rules outside home; 3 = Allowed to smoke in designated areas at
home; 4 = Allowed to smoke anywhere; 5 = I don't set rules about smoking
cigarettes; 6 = I haven't made rules yet about my child smoking
cigarettes, 777 = Decline to answer

\*su_p_rule\_\_nic\_\_cig_001\_\_02\* (formerly parent_rules_q6): Do you
enforce penalties for violating family rules about smoking? 0 = Never; 1
= Rarely; 2 = Sometimes; 3 = Often; 4 = I have not had to enforce these
rules, 777 = decline to answer

Marijuana

\*su_p_rule\_\_mj_001\* (formerly parent_rules_q7): What are the family
rules about using marijuana for your child? 1 = Not allowed to use
marijuana under any circumstances; 2 = Not allowed to use marijuana in
home - no family rules outside home; 3 = Allowed to use marijuana in
home with permission; 4 = Allowed to use marijuana in home whenever
he/she wants; 5 = I don't set rules about marijuana use; 6 = I haven't
made rules yet about my child's marijuana use, 777 = decline to answer

\*su_p_rule\_\_mj_001\_\_02\* (formerly parent_rules_q9) Do you enforce
penalties for violating family rules about using marijuana? 1 = Never; 2
= Rarely; 3 = Sometimes; 4 = Often; 5 = I have not had to enforce these
rules, 777 = decline to answer

## Data Cleaning and Variable Creation

```{r parmon, message=FALSE}

#Subset substance use rules dataset to variables of interest
s.surule <- c("participant_id", "session_id", "su_p_rule__alc_001", "su_p_rule__alc_001__03", "su_p_rule__nic__cig_001", "su_p_rule__nic__cig_001__02", "su_p_rule__mj_001", "su_p_rule__mj_001__02")
surule_subset <- surule[s.surule]

#Subset parental monitoring dataset to variables of interest
s.pm <- c("participant_id", "session_id", "fc_y_pm_mean")
pm_subset <- pm[s.pm]

#Subset family cohesion and conflict dataset to variables of interest
s.fes <- c("participant_id", "session_id","fc_y_fes__cohes_mean","fc_y_fes__confl_mean")
fes_subset <- fes[s.fes]

#Subset peer behavior dataset to variables of interest
s.pbp <- c("participant_id","session_id","fc_y_pbp_001","fc_y_pbp_002","fc_y_pbp_003","fc_y_pbp_004","fc_y_pbp_005","fc_y_pbp_006")
pbp_subset <- pbp[s.pbp]
pbp_subset <- pbp_subset %>%
  mutate(across(c(fc_y_pbp_001,fc_y_pbp_002,fc_y_pbp_003,fc_y_pbp_004,fc_y_pbp_005,fc_y_pbp_006),~ ifelse(.x == 999, NA, as.numeric(.x)))
  ) %>%
  mutate(negative_peer_mean = rowMeans(across(c(fc_y_pbp_001, fc_y_pbp_002, fc_y_pbp_003)),na.rm = TRUE),positive_peer_mean = rowMeans(across(c(fc_y_pbp_004, fc_y_pbp_005, fc_y_pbp_006)),na.rm = TRUE),negative_peer_mean = ifelse(is.nan(negative_peer_mean), NA, negative_peer_mean),positive_peer_mean = ifelse(is.nan(positive_peer_mean), NA, positive_peer_mean))
pbp_subset <- pbp_subset %>%
  select(participant_id, session_id,negative_peer_mean,positive_peer_mean)

#Subset resistance to peer influence dataset to variables of interest
s.rpi <- c("participant_id", "session_id","fc_y_rpi_mean")
rpi_subset <- rpi[s.rpi]

#pm_subset and surule_subset are measured on the annual assessments. However, sdev_subset is a substudy offered at times that may or may not correspond to the annual assessment. 
#Merge pm and surule together for data cleaning and variable creation and add cleaned sdev later. 
parent_peer_datavars <- list(pm_subset, surule_subset, fes_subset, pbp_subset, rpi_subset)
parmon <- Reduce(function(x, y) merge(x, y, by = c("participant_id", "session_id"), all = TRUE), parent_peer_datavars)

#The parental monitoring and substance use rules variables are factors, not numeric, which will mess up zscores later. Recode as numeric
parmon$su_p_rule__alc_001__03n <- as.numeric(as.character(parmon$su_p_rule__alc_001__03))
parmon$su_p_rule__alc_001n <- as.numeric(as.character(parmon$su_p_rule__alc_001))
parmon$su_p_rule__nic__cig_001__02n <- as.numeric(as.character(parmon$su_p_rule__nic__cig_001__02))
parmon$su_p_rule__nic__cig_001n <- as.numeric(as.character(parmon$su_p_rule__nic__cig_001))
parmon$su_p_rule__mj_001__02n <- as.numeric(as.character(parmon$su_p_rule__mj_001__02))
parmon$su_p_rule__mj_001n <- as.numeric(as.character(parmon$su_p_rule__mj_001))

#recode parent substance use rules so that 4 = I have not had to enforce these rules is 0 and so yes= 3 to match tobacco and marijuana.
#alcohol
parmon  <- parmon  %>% 
  mutate(alcrulesenforce = case_when(
    (su_p_rule__alc_001__03n == 0 | su_p_rule__alc_001__03n ==4) ~ 0, 
    (su_p_rule__alc_001__03n == 1) ~ 3,
    TRUE ~ NA)) 

#reverse code substance use rules so that higher numbers indicate stricter rules.
parmon  <- parmon  %>% 
  mutate(alcrules = case_when(
    (su_p_rule__alc_001n == 5 | su_p_rule__alc_001n == 6) ~ 0, 
    (su_p_rule__alc_001n == 4) ~ 1,
    (su_p_rule__alc_001n == 3) ~ 2,
    (su_p_rule__alc_001n == 2) ~ 3,
    (su_p_rule__alc_001n == 1) ~ 4,
    TRUE ~ NA)) 

#tobacco
parmon  <- parmon  %>% 
  mutate(tobrulesenforce = case_when(
    (su_p_rule__nic__cig_001__02n == 0 | su_p_rule__nic__cig_001__02n == 4) ~ 0, 
    (su_p_rule__nic__cig_001__02n == 1) ~ 1,
    (su_p_rule__nic__cig_001__02n == 2) ~ 2,
    (su_p_rule__nic__cig_001__02n == 3) ~ 3,
    TRUE ~ NA)) 

#reverse code substance use rules so that higher numbers indicate stricter rules.
parmon  <- parmon  %>% 
  mutate(tobrules = case_when(
    (su_p_rule__nic__cig_001n == 6 | su_p_rule__nic__cig_001n ==5 ) ~ 0, 
    (su_p_rule__nic__cig_001n == 4) ~ 1,
    (su_p_rule__nic__cig_001n == 3) ~ 2,
    (su_p_rule__nic__cig_001n == 2) ~ 3,
    (su_p_rule__nic__cig_001n == 1) ~ 4,
    TRUE ~ NA))

#marijuana
parmon  <- parmon  %>% 
  mutate(mjrulesenforce = case_when(
    (su_p_rule__mj_001__02n == 1 | su_p_rule__mj_001__02n ==5) ~ 0, 
    (su_p_rule__mj_001__02n == 2) ~ 1,
    (su_p_rule__mj_001__02n == 3) ~ 2,
    (su_p_rule__mj_001__02n == 4) ~ 3,
    TRUE ~ NA)) 

#reverse code substance use rules so that higher numbers indicate stricter rules.
parmon  <- parmon %>% 
  mutate(mjrules = case_when(
    (su_p_rule__mj_001 == 6 | su_p_rule__mj_001 ==5 ) ~ 0, 
    (su_p_rule__mj_001 == 4) ~ 1,
    (su_p_rule__mj_001 == 3) ~ 2,
    (su_p_rule__mj_001 == 2) ~ 3,
    (su_p_rule__mj_001 == 1) ~ 4,
    TRUE ~ NA)) 

#Standardize variables prior to subscale creation
#create formula to generate z-scores
calculate_z <- function(X, X_mean, S){
  return((X-X_mean)/S)
}

parmon <- parmon %>% 
  rename(ch_gen_know = fc_y_pm_mean)

#Child Perceived General Monitoring / Knowledge
parmon <- parmon %>% group_by(session_id) %>% 
  mutate(zch_gen_know = calculate_z(ch_gen_know, 
      mean(ch_gen_know, na.rm = TRUE),
      sd(ch_gen_know, na.rm = TRUE)))

##Parental Control Variables
parmon <- parmon  %>% group_by(session_id) %>% 
  mutate(zsurq1 = calculate_z(alcrules, 
      mean(alcrules, na.rm = TRUE),
      sd(alcrules, na.rm = TRUE)))

parmon <- parmon  %>% group_by(session_id) %>% 
  mutate(zsurq3 = calculate_z(alcrulesenforce, 
      mean(alcrulesenforce, na.rm = TRUE),
      sd(alcrulesenforce, na.rm = TRUE)))

parmon  <- parmon  %>% group_by(session_id) %>% 
  mutate(zsurq4 = calculate_z(tobrules, 
      mean(tobrules, na.rm = TRUE),
      sd(tobrules, na.rm = TRUE)))

parmon  <- parmon  %>% group_by(session_id) %>% 
  mutate(zsurq6 = calculate_z(tobrulesenforce, 
      mean(tobrulesenforce, na.rm = TRUE),
      sd(tobrulesenforce, na.rm = TRUE)))

parmon <- parmon %>% group_by(session_id) %>% 
  mutate(zsurq7 = calculate_z(mjrules, 
      mean(mjrules, na.rm = TRUE),
      sd(mjrules, na.rm = TRUE)))

parmon  <- parmon  %>% group_by(session_id) %>% 
  mutate(zsurq9 = calculate_z(mjrulesenforce, 
      mean(mjrulesenforce, na.rm = TRUE),
      sd(mjrulesenforce, na.rm = TRUE)))

parmon <- parmon %>% rowwise() %>%
    mutate(par_su_rules= mean(c(zsurq1, zsurq3, zsurq4, zsurq6, zsurq7, zsurq9), na.rm=TRUE))

#pairdown to just outcome variables
s.pm <- c("participant_id", "session_id", "ch_gen_know", "par_su_rules", "fc_y_fes__cohes_mean", "fc_y_fes__confl_mean", "negative_peer_mean", "positive_peer_mean", "fc_y_rpi_mean")

parmon <- parmon[s.pm]

#Note: Final data frame name parmon

```

## Viz: Distributions across all waves

```{r, echo=FALSE, warning=FALSE, message=FALSE}

plot_distribution(
  data = parmon,
  var = ch_gen_know,
  title = "Child-Perceived Parental Monitoring",
  xlab = "Parental Monitoring",
  annotate_x = 3
)

plot_distribution(
  data = parmon,
  var = par_su_rules,
  title = "Parent Substance Use Rules",
  xlab = "Parent Substance Use Rules",
  annotate_x = 1
)

plot_distribution(
  data = parmon,
  var = fc_y_fes__cohes_mean,
  title = "Child-Perceived Family Cohesion",
  xlab = "Family Cohesion",
  annotate_x = 1.5
)

plot_distribution(
  data = parmon,
  var = fc_y_fes__confl_mean,
  title = "Child-Perceived Family Conflict",
  xlab = "Family Conflict",
  annotate_x = .5
)

plot_distribution(
  data = parmon,
  var = negative_peer_mean,
  title = "Negative Peer Characteristics",
  xlab = "Negative Peer Characteristics",
  annotate_x = 2.2
)

plot_distribution(
  data = parmon,
  var = positive_peer_mean,
  title = "Positive Peer Characteristics",
  xlab = "Positive Peer Characteristics",
  annotate_x = 1
)

plot_distribution(
  data = parmon,
  var = fc_y_rpi_mean,
  title = "Resistance to Peer Influence",
  xlab = "Resistance to Peer Influence",
  annotate_x = 2
)


```

## Viz: Distributions by wave

```{r, echo=FALSE, warning=FALSE, message=FALSE}

plot_distribution_by_time(
  data = parmon,
  var = ch_gen_know,
  time_var = session_id,
  title = "Child-Perceived Parental Monitoring by Wave",
  xlab = "Parental Monitoring",
  annotate_x = 3
)

plot_distribution_by_time(
  data = parmon,
  var = par_su_rules,
  time_var = session_id,
  title = "Parent Substance Use Rules by Wave",
  xlab = "Parent Substance Use Rules",
  annotate_x = 1
)

plot_distribution_by_time(
  data = parmon,
  var = fc_y_fes__cohes_mean,
  time_var = session_id,
  title = "Child-Perceived Family Cohesion by Wave",
  xlab = "Family Cohesion",
  annotate_x = 1.5
)

plot_distribution_by_time(
  data = parmon,
  var = fc_y_fes__confl_mean,
  time_var = session_id,
  title = "Child-Perceived Family Conflict by Wave",
  xlab = "Family Conflict",
  annotate_x = .5
)

plot_distribution_by_time(
  data = parmon,
  var = negative_peer_mean,
  time_var = session_id,
  title = "Negative Peer Characteristics by Wave",
  xlab = "Negative Peer Characteristics",
  annotate_x = 2.2
)

plot_distribution_by_time(
  data = parmon,
  var = positive_peer_mean,
  time_var = session_id,
  title = "Positive Peer Characteristics by Wave",
  xlab = "Positive Peer Characteristics",
  annotate_x = 1
)

plot_distribution_by_time(
  data = parmon,
  var = fc_y_rpi_mean,
  time_var = session_id,
  title = "Resistance to Peer Influence by Wave",
  xlab = "Resistance to Peer Influence",
  annotate_x = 2
)
```

## Viz: Longitudinal Plots

```{r, echo=FALSE, warning=FALSE, message=FALSE}

parmon <- parmon %>% 
  mutate(timepoint = str_extract(session_id, "\\d+") %>% as.integer())

plot_longitudinal_subset(
  data = parmon,
  var = ch_gen_know,
  time_var = timepoint,
  id_var = participant_id,
  sample_n = 50,
  seed = 2025,
  title = "Child-Perceived Parental Monitoring over Time",
  xlab = "Time",
  ylab = "Parental Monitoring"
)

plot_longitudinal_subset(
  data = parmon,
  var = par_su_rules,
  time_var = timepoint,
  id_var = participant_id,
  sample_n = 50,
  seed = 2025,
  title = "Parent Substance Use Rules over Time",
  xlab = "Time",
  ylab = "Parent Substance Use Rules"
)

plot_longitudinal_subset(
  data = parmon,
  var = fc_y_fes__cohes_mean,
  time_var = timepoint,
  id_var = participant_id,
  sample_n = 50,
  seed = 2025,
  title = "Child-Perceived Family Cohesion over Time",
  xlab = "Time",
  ylab = "Family Cohesion"
)

plot_longitudinal_subset(
  data = parmon,
  var = fc_y_fes__confl_mean,
  time_var = timepoint,
  id_var = participant_id,
  sample_n = 50,
  seed = 2025,
  title = "Child-Perceived Family Conflict over Time",
  xlab = "Time",
  ylab = "Family Conflict"
)

plot_longitudinal_subset(
  data = parmon,
  var = negative_peer_mean,
  time_var = timepoint,
  id_var = participant_id,
  sample_n = 50,
  seed = 2025,
  title = "Negative Peer Characteristics over Time",
  xlab = "Time",
  ylab = "Negative Peer Characteristics"
)

plot_longitudinal_subset(
  data = parmon,
  var = positive_peer_mean,
  time_var = timepoint,
  id_var = participant_id,
  sample_n = 50,
  seed = 2025,
  title = "Positive Peer Characteristics over Time",
  xlab = "Time",
  ylab = "Positive Peer Characteristics"
)

plot_longitudinal_subset(
  data = parmon,
  var = fc_y_rpi_mean,
  time_var = timepoint,
  id_var = participant_id,
  sample_n = 50,
  seed = 2025,
  title = "Resistance to Peer Influence over Time",
  xlab = "Time",
  ylab = "Resistance to Peer Influence"
)
```


# Linked External Data

## Dataset(s) & Variables

**le_l_adi**: Area Deprivation Index; a neighborhood-level composite
measure of socioeconomic disadvantage. Higher values indicate greater
neighborhood deprivation. variables were measured at baseline (00A).

*le_l_adi\_\_addr1\_\_income_med*: Median family income [Area
Deprivation Index] (Address 1)

*le_l_adi\_\_addr1\_\_mortgage_med*: Median monthly mortgage [Area
Deprivation Index] (Address 1)

**le_l_svi**: Social Vulnerability Index; a neighborhood-level composite
measure of community vulnerability based on social, economic,
demographic, and housing-related factors. Higher values indicate greater
social vulnerability. variables were measured at baseline (00A)

*le_l_svi\_\_addr1\_\_17younger_prcnt*: Percent population estimate of persons 17 and younger subcomponent [CDC SVI (ACS 2014-2018 5 yr avg); census tract] (Address 1)

**le_l_densalc**: Alcohol outlet density based on Census 2016 measures;
captures the density of alcohol outlets in the participant's residential
area. Higher values indicate greater alcohol outlet density. variables were measured at baseline (00A)

*le_l_densalc\_\_addr1\_\_alc_density\_\_2016*: Estimate of alcohol
outlet density (Number of Establishments where alcohol could be
purchased and/or consumed based on North American Industry
Classification System (NAICS) from ZIP Code Business Patterns from Census 2016) (Address 1)

## Data Cleaning & Variable Creation

```{r ed, , message=FALSE}

#Subset brief adi/svi/densalc to variables of interest
s.adi <- c("participant_id", "session_id","le_l_adi__addr1__income_med", "le_l_adi__addr1__mortgage_med")
adi_subset <- adi[s.adi]

s.svi <- c("participant_id", "session_id","le_l_svi__addr1__17younger_prcnt")
svi_subset <- svi[s.svi]

s.densalc <- c( "participant_id", "session_id","le_l_densalc__addr1__alc_density__2016")
densalc_subset <- densalc[s.densalc]

le <- adi_subset %>%
  full_join(svi_subset, by = c("participant_id", "session_id")) %>%
  full_join(densalc_subset, by = c("participant_id", "session_id"))

# Final data frame name le
```

## Viz: Distributions of linked external data at baseline

```{r, echo=FALSE, warning=FALSE, message=FALSE}

plot_distribution(
  data = le,
  var = le_l_adi__addr1__income_med,
  title = "Median Family Income",
  xlab = "Median Family Income",
  annotate_x = 100200
)

plot_distribution(
  data = le,
  var = le_l_adi__addr1__mortgage_med,
  title = "Median Monthly Mortgage",
  xlab = "Median Monthly Mortgage",
  annotate_x = 2000
)

plot_distribution(
  data = le,
  var = le_l_svi__addr1__17younger_prcnt,
  title = "Percent Population Aged 17 or Younger",
  xlab = "Percent Population Aged 17 or Younger",
  annotate_x = 30
)

plot_distribution(
  data = le,
  var = le_l_densalc__addr1__alc_density__2016,
  title = "Alcohol Outlet Density Estimate, 2016",
  xlab = "Alcohol Outlet Density Estimate",
  annotate_x = 150
)
```

# Twin/Sibling/Family Data

We used the twin/sibling subset of the Adolescent Brain and Cognitive
Development Study.

Sibling/Twin Identifiers:

## Dataset(s) & Variables Used

**gn_e_zygrat**

*gn_e_zygrat_famno* (formerly zyg_ss_fam_no in gen_y_zygrat): Family
Number. *gn_e_zygrat_judgement* (formerly zyg_ss_r1_judgement in
gen_y_zygrat): Rating. (1 = DZ; 2 = MZ)

**gn_y_genrel**

\*gn_y_genrel_zyg\_\_01 through gn_y_genrel_zyg\_\_04\* (formerly
genetic_zygosity_status\_(1,2,3,or 4)): Genetically inferred zygosity
status between participant and gn_y_genrel_id\_\_paired\_\_01 through
gn_y_genrel_id\_\_paired\_\_04. 1 = Monozygotic, 2 = Dizygotic, 3 =
sibling, 444 = not available \*gn_y_genrel_id\_\_paired\_\_01 through
gn_y_genrel_id\_\_paired\_\_04\* (formerly
genetic_paired_subjectid\_(1,2,3, or 4)): ID of the participant who is
genetic related based on gn_y_genrel_pihat\_\_01 through
gn_y_genrel_pihat\_\_04

## Data Cleaning & Variable Creation

```{r gendata, message=FALSE}


#Subset zygrat dataset to variables of interest
s.zygrat <- c("participant_id", "session_id", "gn_e_zygrat_famno", "gn_e_zygrat_judgement")
zygrat_subset <- zygrat[s.zygrat]

#Subset genrel dataset to variables of interest
s.genrel <- c("participant_id", "gn_y_genrel_zyg__01", "gn_y_genrel_zyg__02", "gn_y_genrel_zyg__03", "gn_y_genrel_zyg__04", "gn_y_genrel_id__paired__01", "gn_y_genrel_id__paired__02", "gn_y_genrel_id__paired__03", "gn_y_genrel_id__paired__04")
genrel_subset <- genrel[s.genrel]


#merge genetic specific dataframes with demo dataframe 
#Note: genrel_subset doesn't have a session_id var. Merge demo_mf with zygrat_subset, then merge combined dataframe with genrel
rel1 <- merge(demo_mf, zygrat_subset, by=c("participant_id", "session_id"), all = TRUE)
rel2 <- merge(rel1 , genrel_subset, by=c("participant_id"), all = TRUE)

#sort the data so it is in time order and families are together
sort.rel <- rel2[order(rel2$ab_g_stc__design_id__fam, rel2$participant_id, rel2$session_id), ]

#gn_y_genrel_zyg vars have 444 when genetic_pi_hat not calculated. Change 444 to NA
sort.rel2 <- sort.rel %>% 
mutate(genrel_zyg01 = if_else(gn_y_genrel_zyg__01 %in% c(444), NA, gn_y_genrel_zyg__01)) %>%
mutate(genrel_zyg02 = if_else(gn_y_genrel_zyg__02 %in% c(444), NA, gn_y_genrel_zyg__02)) %>%
mutate(genrel_zyg03 = if_else(gn_y_genrel_zyg__03 %in% c(444), NA, gn_y_genrel_zyg__03)) %>%
mutate(genrel_zyg04 = if_else(gn_y_genrel_zyg__04 %in% c(444), NA, gn_y_genrel_zyg__04))

class(sort.rel2$genrel_zyg01)
#Note: Although the gn_y_genrel_zyg variables contain numeric values, it is a factor. Change to numeric for future variable manipulation

sort.rel2$genrel_zyg01n <- as.numeric(as.character(sort.rel2$genrel_zyg01))
sort.rel2$genrel_zyg02n <- as.numeric(as.character(sort.rel2$genrel_zyg02))
sort.rel2$genrel_zyg03n <- as.numeric(as.character(sort.rel2$genrel_zyg03))
sort.rel2$genrel_zyg04n <- as.numeric(as.character(sort.rel2$genrel_zyg04))

class(sort.rel2$genrel_zyg01n)

#Combine genetic zygosity variables into a single variable in which the lowest zygosity (i.e., 1=MZ, 2=DZ, 3=siblings) is retained. For example, individuals with famid 1430 has a set of twins and their sibling. This code will maintain the twin zygosity status instead of the 3 they share with their genetically related sibling.  
sort.rel2 <- sort.rel2 %>%
  mutate(zygcombined = pmin(genrel_zyg01n, genrel_zyg02n, genrel_zyg03n, genrel_zyg04n, na.rm = TRUE))

#For those who do not have genetic zygosity, will use interviewer determined zygosity. 
#first, we need to ensure that zygosity judgement matches for the twin pairs because sometimes they get different results. For example, twins with famid 11878, one twin received a judgement of 1 (DZ) while the other received a judgement of 2 (MZ). For twins without genetic zygosity whose judgement don't match, we will set them as siblings. 

sort.rel2 <- sort.rel2 %>% 
group_by(ab_g_stc__design_id__fam, session_id) %>% 
mutate(consistent_zyg = length(unique(gn_e_zygrat_judgement)) == 1)

table(sort.rel2$consistent_zyg)
#shows there is some inconsistency in matching (i.e., 399 instances of FALSE)

#make numeric variable where 1=mismatched and 0=matched
sort.rel2 <- sort.rel2 %>% 
  mutate(zygmismatch = case_when(
    (consistent_zyg == "TRUE") ~ 0, 
    (consistent_zyg == "FALSE") ~ 1,
    TRUE ~ NA))

table(sort.rel2$zygmismatch)
#Note: 399 have a 1 which matches the factor variable

#Make a new dataset to make sure that those marked as mismatched are actually mismatched 
check <- sort.rel2 %>% group_by(ab_g_stc__design_id__fam, session_id) %>%
filter(zygmismatch==1)

#check on those families who have more than 2 people in them
check_morethan2 <- sort.rel2 %>% 
  group_by(ab_g_stc__design_id__fam, session_id) %>% 
  mutate(count = n_distinct(participant_id)) %>% 
  filter(zygmismatch == 1 & count > 2)

#recode judgement zygosity coding so it matches genetic zygosity wherein 1=MZ, 2=DZ
sort.rel2 <- sort.rel2 %>% 
  mutate(judgezygr = case_when(
    (gn_e_zygrat_judgement == 1) ~ 2, 
    (gn_e_zygrat_judgement == 2) ~ 1,
    TRUE ~ NA))

sort.rel3 <- sort.rel2 %>% ungroup() %>%
mutate (zyg = if_else(!is.na(zygcombined), zygcombined, ## If zygcombined isn't missing, take zygcombined
      if_else(zygmismatch == 1, ## if zygcombined is missing, check mismatch 
         3, ## If they are mismatched, mark as sibling (3)
         judgezygr)))  ## Otherwise, put judgezygr 

#remove the same demo variables
rel <- sort.rel3 %>%
  select(
    -ab_g_dyn__visit_dtt,
    -ab_g_dyn__visit_age,
    -ab_p_demo__relation_001,
    -ab_p_demo__marital__slf_001,
    -multigen,
    -ab_p_demo__child__time_001,
    -ab_p_demo__child__time_001__01,
    -ab_p_demo__child__time_001__02,
    -ab_g_stc__cohort_ethnrace__leg,
    -ab_g_stc__cohort_race__nih,
    -ab_g_stc__cohort_sex,
    -ab_g_stc__design_id__fam,
    -ab_g_stc__design_id__birth,
    -ab_g_stc__design_famrel,
    -housepop
  )

#Note: Final data frame name rel

```

# Combing all finalized datasets

```{r, message=FALSE}

# List of datasets to merge 
all_final_datasets <- list(demo_mf, rel, mh, su, parmon, le)

# Merge all data.tables using full outer join based on id
all_final_merged <- Reduce(function(x, y) merge(x, y, by = c("participant_id", "session_id"), all = TRUE), all_final_datasets)

sort.all_final_merged <- all_final_merged[order(all_final_merged$participant_id, all_final_merged$session_id), ]

sort.all_final_merged <- sort.all_final_merged %>%
  select(-timepoint, -timepoint.y, -timepoint.x, -ab_g_dyn__design_site.y) %>%
  rename(ab_g_dyn__design_site=ab_g_dyn__design_site.x)

sort.all_final_merged <- sort.all_final_merged %>%
  mutate(timepoint = str_extract(session_id, "\\d+") %>% as.integer())

#Save out final merged dataset
write.csv(sort.all_final_merged, "C:\\Users\\kmarcea\\Box\\MCAP Steering Committee\\Summer Institute\\- Summer Institute - Shared Folder\\Datasets\\MCAPSI_ABCDv7_2026_07_01.csv", row.names = FALSE)

#write.csv(sort.all_final_merged,"C:/Users/admin/Box/- Summer Institute - Shared Folder/Datasets/MCAPSI_ABCDv7_2026_06_02.csv",row.names = FALSE)

sort.all_final_merged %>%
  summarise(
    n_families = n_distinct(ab_g_stc__design_id__fam),
    n_people = n_distinct(participant_id),
    max_rows_per_person = max(table(participant_id))
  )

# Select one participant per family
set.seed(123)

one_person_per_family <- sort.all_final_merged %>%
  filter(
    !is.na(ab_g_stc__design_id__fam),
    !is.na(participant_id)
  ) %>%
  distinct(ab_g_stc__design_id__fam, participant_id) %>%
  group_by(ab_g_stc__design_id__fam) %>%
  slice_sample(n = 1) %>%
  ungroup()

ABCD_onedata <- sort.all_final_merged %>%
  semi_join(
    one_person_per_family,
    by = c("ab_g_stc__design_id__fam", "participant_id")
  )

ABCD_onedata %>%
  summarise(
    n_families = n_distinct(ab_g_stc__design_id__fam),
    n_people = n_distinct(participant_id),
    max_rows_per_person = max(table(participant_id))
  )

write.csv(ABCD_onedata, "C:\\Users\\kmarcea\\Box\\MCAP Steering Committee\\Summer Institute\\- Summer Institute - Shared Folder\\Datasets\\MCAPSI_ABCDv7_2026_06_08_oneperfam.csv", row.names = FALSE)

#write.csv(ABCD_onedata, "C:/Users/admin/Box/- Summer Institute - Shared Folder/Datasets/MCAPSI_ABCDv7_2026_07_07_oneperfam.csv", row.names = FALSE)

```

# Alphabetical Variable Dictionary

## A

```{r, echo=FALSE, warning=FALSE, message=FALSE}
## data.frame to contain info; putting it in a data.frame allows the table to be prettified with kableExtra
A <- tribble(
  ~ indicator, ~ description,
"`ab_g_dyn__visit_dtt`", "Date and time at the start of the visit.",
  "`ab_g_dyn__visit_age`", "Youth age at the start of the visit.",
  "`ab_p_demo__relation_001`", "Relationship of the reporting caregiver to the participant.",
  "`ab_p_demo__marital__slf_001`", "Marital status of the reporting caregiver.",
"`ab_p_demo__child__time_001`", "Whether the child spends significant time in another household.",
  "`ab_p_demo__child__time_001__01`", "Amount of time per week the child spends in another household.",
  "`ab_p_demo__child__time_001__02`", "Best person to contact about the other household.",
  "`ab_g_stc__cohort_ethnrace__leg`", "Parent-reported child race/ethnicity. 1 = Hispanic; 2 = White; 3 = Black; 4 = Asian; 13 = Other.",
  "`ab_g_stc__cohort_race__nih`", "Child race based on NIH classification.",
  "`ab_g_stc__cohort_sex`", "Parent-reported sex assigned at birth. 1 = Male; 2 = Female.",
  "`ab_g_stc__design_id__fam`", "Family ID. Participants from the same family share a family ID.",
  "`ab_g_stc__design_id__birth`", "Birth event ID. Participants born together share a birth event ID.",
  "`ab_g_stc__design_famrel`", "Family relationship structure indicator.",
 "`AUC`", "Alcohol uptake continuum. Higher scores indicate greater alcohol use involvement.",
"`AU_init`", "Alcohol use initiation indicator. 0 = no initiation; 1 = alcohol uptake score greater than or equal to 1."
)

```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
kable(
  A,
  format = "html",
  col.names = c("Variable Name", "Variable Description")
) %>%
  kable_styling(bootstrap_options = c("hover"))
```

## C

```{r, echo=FALSE, warning=FALSE, message=FALSE}
## data.frame to contain info; putting it in a data.frame allows the table to be prettified with kableExtra
C <- tribble(
  ~ indicator, ~ description,
  "`ch_gen_know`", "Child-perceived parental monitoring/general knowledge, based on the youth-reported parental monitoring mean score.",
  "`consistent_zyg`", "Indicator for whether interviewer-rated zygosity was consistent within family and session."
)

```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
kable(
  C,
  format = "html",
  col.names = c("Variable Name", "Variable Description")
) %>%
  kable_styling(bootstrap_options = c("hover"))
```

## F

```{r, echo=FALSE, warning=FALSE, message=FALSE}
## data.frame to contain info; putting it in a data.frame allows the table to be prettified with kableExtra
f <- tribble(
  ~ indicator, ~ description,
  "`fc_y_fes__cohes_mean`", "Youth-reported Family Environment Scale cohesion mean score.",
  "`fc_y_fes__confl_mean`", "Youth-reported Family Environment Scale conflict mean score.",
  "`fc_y_rpi_mean`", "Youth-reported Resistance to Peer Influence mean score. Higher scores indicate greater resistance to peer influence."
)

```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
kable(
  f,
  format = "html",
  col.names = c("Variable Name", "Variable Description")
) %>%
  kable_styling(bootstrap_options = c("hover"))
```

## G

```{r, echo=FALSE, warning=FALSE, message=FALSE}
## data.frame to contain info; putting it in a data.frame allows the table to be prettified with kableExtra
g <- tribble(
  ~ indicator, ~ description,
  "`gn_e_zygrat_famno`", "Family number from the zygosity rating dataset.",
  "`gn_e_zygrat_judgement`", "Interviewer-rated twin zygosity. Original coding: 1 = DZ; 2 = MZ.",
  "`gn_y_genrel_zyg__01`", "Genetically inferred zygosity status for paired participant 1. 1 = MZ; 2 = DZ; 3 = sibling; 444 = not available.",
  "`gn_y_genrel_zyg__02`", "Genetically inferred zygosity status for paired participant 2. 1 = MZ; 2 = DZ; 3 = sibling; 444 = not available.",
  "`gn_y_genrel_zyg__03`", "Genetically inferred zygosity status for paired participant 3. 1 = MZ; 2 = DZ; 3 = sibling; 444 = not available.",
  "`gn_y_genrel_zyg__04`", "Genetically inferred zygosity status for paired participant 4. 1 = MZ; 2 = DZ; 3 = sibling; 444 = not available.",
  "`gn_y_genrel_id__paired__01`", "Participant ID of genetically related paired participant 1.",
  "`gn_y_genrel_id__paired__02`", "Participant ID of genetically related paired participant 2.",
  "`gn_y_genrel_id__paired__03`", "Participant ID of genetically related paired participant 3.",
  "`gn_y_genrel_id__paired__04`", "Participant ID of genetically related paired participant 4.",
)

```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
kable(
  g,
  format = "html",
  col.names = c("Variable Name", "Variable Description")
) %>%
  kable_styling(bootstrap_options = c("hover"))
```

## H

```{r, echo=FALSE, warning=FALSE, message=FALSE}
## data.frame to contain info; putting it in a data.frame allows the table to be prettified with kableExtra
h <- tribble(
  ~ indicator, ~ description,
    "`housepop`", "Number of people living at the youth's address."
)

```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
kable(
  h,
  format = "html",
  col.names = c("Variable Name", "Variable Description")
) %>%
  kable_styling(bootstrap_options = c("hover"))
```

## J

```{r, echo=FALSE, warning=FALSE, message=FALSE}
## data.frame to contain info; putting it in a data.frame allows the table to be prettified with kableExtra
j <- tribble(
  ~ indicator, ~ description,
    "`judgezygr`", "Recoded interviewer-rated zygosity to match genetic zygosity coding. 1 = MZ; 2 = DZ."
)

```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
kable(
  j,
  format = "html",
  col.names = c("Variable Name", "Variable Description")
) %>%
  kable_styling(bootstrap_options = c("hover"))
```


## L

```{r, echo=FALSE, warning=FALSE, message=FALSE}
## data.frame to contain info; putting it in a data.frame allows the table to be prettified with kableExtra
l <- tribble(
  ~ indicator, ~ description,
    "`le_l_adi__addr1__income_med`", 
  "Median family income used as a subcomponent of the Area Deprivation Index, based on Address 1. Measured at baseline (00A).",
  
  "`le_l_adi__addr1__mortgage_med`", 
  "Median monthly mortgage used as a subcomponent of the Area Deprivation Index, based on Address 1. Measured at baseline (00A).",
  
  "`le_l_svi__addr1__17younger_prcnt`", 
  "Percent population estimate of persons aged 17 years or younger, used as a subcomponent of the CDC Social Vulnerability Index, based on census tract information from Address 1. Measured at baseline (00A).",
  
  "`le_l_densalc__addr1__alc_density__2016`", 
  "Estimated alcohol outlet density, defined as the number of establishments where alcohol could be purchased and/or consumed, based on NAICS classifications from ZIP Code Business Patterns from Census 2016. Measured at baseline (00A)."
)

```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
kable(
  l,
  format = "html",
  col.names = c("Variable Name", "Variable Description")
) %>%
  kable_styling(bootstrap_options = c("hover"))
```

## M

```{r, echo=FALSE, warning=FALSE, message=FALSE}
## data.frame to contain info; putting it in a data.frame allows the table to be prettified with kableExtra
m <- tribble(
  ~ indicator, ~ description,
 "`mh_p_cbcl__synd__ext_sum`", "Parent-reported CBCL externalizing syndrome scale sum.",
  "`mh_p_cbcl__synd__int_sum`", "Parent-reported CBCL internalizing syndrome scale sum.",
  "`mh_p_cbcl__synd__attn_sum`", "Parent-reported CBCL attention problems syndrome scale sum.",
  "`mh_y_bpm__ext_sum`", "Youth-reported Brief Problem Monitor externalizing sum score.",
  "`mh_y_bpm__int_sum`", "Youth-reported Brief Problem Monitor internalizing sum score.",
  "`mh_y_pai_sum`", "Youth-reported NIH Toolbox Positive Affect Items sum score.",
  "`multigen`", "Indicator for whether the youth lives in a multigenerational household.",
  "`MUC`", "Marijuana uptake continuum. Higher scores indicate greater marijuana use involvement.",
  "`MU_init`", "Marijuana use initiation indicator. 0 = no initiation; 1 = marijuana uptake score greater than or equal to 1."
)

```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
kable(
  m,
  format = "html",
  col.names = c("Variable Name", "Variable Description")
) %>%
  kable_styling(bootstrap_options = c("hover"))
```

## N

```{r, echo=FALSE, warning=FALSE, message=FALSE}
## data.frame to contain info; putting it in a data.frame allows the table to be prettified with kableExtra
n <- tribble(
  ~ indicator, ~ description,
 "`negative_peer_mean`", "Mean of negative peer behavior items, including friends skipping school, being suspended, and shoplifting."
)

```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
kable(
  n,
  format = "html",
  col.names = c("Variable Name", "Variable Description")
) %>%
  kable_styling(bootstrap_options = c("hover"))
```

## P

```{r, echo=FALSE, warning=FALSE, message=FALSE}
## data.frame to contain info; putting it in a data.frame allows the table to be prettified with kableExtra
p <- tribble(
  ~ indicator, ~ description,
 "`participant_id`", "Unique participant identifier.",
  "`par_su_rules`", "Parent-reported substance use rules composite. Higher scores indicate stricter and more strongly enforced rules regarding alcohol, tobacco, and marijuana.",
 "`positive_peer_mean`", "Mean of positive peer characteristic items, including friends being athletes, attending religious/spiritual services, and being excellent students."

)

```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
kable(
  p,
  format = "html",
  col.names = c("Variable Name", "Variable Description")
) %>%
  kable_styling(bootstrap_options = c("hover"))
```

## S

```{r, echo=FALSE, warning=FALSE, message=FALSE}
## data.frame to contain info; putting it in a data.frame allows the table to be prettified with kableExtra
s <- tribble(
  ~ indicator, ~ description,
 "`session_id`", "ABCD assessment session identifier.",
 "`SUC`", "Overall substance use continuum/involvement score, calculated from alcohol, tobacco, and marijuana uptake continua.",
  "`SU_init`", "Any substance use initiation indicator. 0 = no substance use initiation; 1 = initiation of alcohol, tobacco, or marijuana use."
)

```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
kable(
  s,
  format = "html",
  col.names = c("Variable Name", "Variable Description")
) %>%
  kable_styling(bootstrap_options = c("hover"))
```

## T

```{r, echo=FALSE, warning=FALSE, message=FALSE}
## data.frame to contain info; putting it in a data.frame allows the table to be prettified with kableExtra
t <- tribble(
  ~ indicator, ~ description,
 "`timepoint`", "Numeric timepoint extracted from session_id.",
 "`TUC`", "Tobacco uptake continuum. Higher scores indicate greater tobacco/nicotine use involvement.",
 "`TU_init`", "Tobacco use initiation indicator. 0 = no initiation; 1 = tobacco uptake score greater than or equal to 1."
)

```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
kable(
  t,
  format = "html",
  col.names = c("Variable Name", "Variable Description")
) %>%
  kable_styling(bootstrap_options = c("hover"))
```

## Z

```{r, echo=FALSE, warning=FALSE, message=FALSE}
## data.frame to contain info; putting it in a data.frame allows the table to be prettified with kableExtra
z <- tribble(
  ~ indicator, ~ description,
  "`zygcombined`", "Combined genetically inferred zygosity variable, retaining the closest genetic relationship across paired participants.",
   "`zygmismatch`", "Indicator for mismatch in interviewer-rated zygosity within family and session. 0 = matched; 1 = mismatched.",
    "`zyg`", "Final zygosity/family relatedness variable. 1 = MZ twin; 2 = DZ twin; 3 = sibling."
)

```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
kable(
  z,
  format = "html",
  col.names = c("Variable Name", "Variable Description")
) %>%
  kable_styling(bootstrap_options = c("hover"))
```
